---
title: "Figure 2: PMC Plot"
output: html_document
date: "2024-08-29"
editor_options: 
  chunk_output_type: console
---


# Figure 2A: GOplot — PMC vs Non-PMC with Upregulated Genes Using GSEA
```{r Figure_2A_PMC_GOplot}
library(openxlsx)
library(enrichplot)
library(clusterProfiler)
library(GOplot)
library(ggsci)

housekeeping_genes <- data.frame(openxlsx::read.xlsx("Housekeep gene set.xlsx", sheet = 1))
top_markers   <- read.csv("Reactome pathways.csv", header = TRUE)
cluster_marker <- read.csv("Figure 2A PMC vs nonPMC DEGs.csv", header = TRUE)
cluster_marker <- cluster_marker[!grepl("^MT-|^MT|\\.|^(RPS|RPL)", cluster_marker$X), ]
cluster_marker <- cluster_marker[!cluster_marker$X %in% housekeeping_genes$Symbol, ]
cluster_marker <- cluster_marker %>%
  filter(avg_log2FC > log2(1) & p_val_adj < 0.05) %>%
  filter(pct.1 > 0.1 | pct.2 > 0.1)

# Load Reactome gene sets
geneset <- read.gmt("Reactome_human.gmt")
geneset <- geneset[geneset$term %in% top_markers$gene, ]

# Extract FDR and FC values
group1 <- cluster_marker
group1$external_gene_name <- group1$X
group1$logFC <- group1$avg_log2FC
group1$FDR   <- group1$p_val_adj
group1 <- group1[, c("external_gene_name", "logFC", "FDR")]
rownames(group1) <- group1$external_gene_name

# Add up/down regulation column: log2FC > log2(1), adj. p < 0.05
group1$g1 <- "normal"
group1$g1[group1$logFC >  log2(1) & group1$FDR < 0.05] <- "up"
group1$g1[group1$logFC < -log2(1) & group1$FDR < 0.05] <- "down"
table(group1$g1)

# Get differentially expressed genes
DEGs <- group1[group1$g1 != "normal", ]
DEGs <- DEGs[order(DEGs$logFC, decreasing = TRUE), ]
genelist <- DEGs$logFC
names(genelist) <- DEGs$external_gene_name

# Run GSEA
enrichplot_1 <- GSEA(genelist, TERM2GENE = geneset, pvalueCutoff = 1,
                     pAdjustMethod = "fdr", minGSSize = 10, maxGSSize = 10000)
p1 <- dotplot(enrichplot_1, showCategory = 30, label_format = 100, color = "pvalue")
p1

# Build GOplot chord data
enrichplot <- data.frame(enrichplot_1)
go <- data.frame(
  Category = enrichplot$ID,
  ID       = enrichplot$ID,
  Term     = enrichplot$Description,
  Genes    = gsub("/", ", ", enrichplot$core_enrichment),
  adj_pval = enrichplot$p.adjust
)
genelist <- data.frame(ID = group1$external_gene_name, logFC = group1$logFC)
row.names(genelist) <- genelist[, 1]
circ <- circle_dat(go, genelist)

termNum <- 10
termNum  <- ifelse(nrow(go) < termNum, nrow(go), termNum)
geneNum  <- nrow(genelist)
chord    <- chord_dat(circ, genelist[1:geneNum, ], go$Term[1:termNum])

# Helper: insert newline after the second space in long pathway names
add_newline_after_second_space <- function(name) {
  spaces <- gregexpr(" ", name)[[1]]
  # If more than two spaces, insert a newline after the second space
  if (length(spaces) > 2) {
    second_space_pos <- spaces[2]
    name <- paste0(
      substr(name, 1, second_space_pos), "\n",
      substr(name, second_space_pos + 1, nchar(name))
    )
  }
  return(name)
}

colnames(chord) <- sapply(colnames(chord), add_newline_after_second_space)

source("saved code/GOChord_update")
my_colors <- colorRampPalette(brewer.pal(12, "Paired"))(9)

p2_A <- update_GOChord(chord,
  space       = 0.125,
  gene.order  = "logFC",
  gene.space  = 0.25,
  gene.size   = 5,
  ribbon.col  = c(my_colors, "antiquewhite1"),
  border.size = 0,
  process.label = 5
)
p2_A
```


# Figure 2E: RT-qPCR Relative Fold Expression (PF4 and PPBP)
```{r Figure_2E_qPCR}
library(readxl)
library(ggpubr)

dat <- data.frame(read_xlsx("Figure 2E qPCR of PF4 and PPBP.xlsx"))
dat_long <- pivot_longer(dat, cols = c(Mono, Mono_PLT),
                         names_to = "Group", values_to = "Expression")
dat_long$Group <- ifelse(dat_long$Group == "Mono",     "in vitro nonPMC",
                  ifelse(dat_long$Group == "Mono_PLT", "in vitro PMC", NA))

p2_E <- ggplot(dat_long, aes(x = Group, y = Expression, fill = Group)) +
  geom_violin(width = 0.5, alpha = 1, color = NA) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 1) +
  geom_point(position = position_jitter(width = 0.01), size = 2, aes(color = Group)) +
  scale_fill_manual(values  = my_color_PMC) +
  scale_color_manual(values = my_color_PMC) +
  facet_wrap(~ genes, scales = "free_y", ncol = 1) +
  stat_compare_means(
    method      = "t.test",
    label       = "p.format",
    comparisons = list(c("in vitro nonPMC", "in vitro PMC")),
    label.y.npc = "top",
    size        = 5
  ) +
  theme_classic(base_size = 16) +
  labs(x = "", y = "Relative Fold Expression") +
  theme(axis.text.x = element_text(angle = -45, vjust = 0, hjust = 0.2)) +
  scale_y_continuous(expand = c(0, 35))
p2_E
```


# Bulk RNA-seq: Load Raw Count Data
```{r Bulk_RNAseq_load_counts}
library(DESeq2)
library(openxlsx)

# Load gene count matrix
countdata <- read.table("Figure 2 Bulk_RNAseq_deSeq2_counts.txt", header = TRUE, row.names = 1)

# Load sample group information
coldata <- read.xlsx("Figure 2 Bulk_RNAseq_Groups.xlsx", rowNames = TRUE)
coldata$group <- c(rep("Mono", 7), rep("Mono_PLT_low", 7), rep("Mono_PLT", 7),
                   rep("Mono", 8), rep("Mono_PLT_low", 8), rep("Mono_PLT", 8))
coldata$day <- "D2"
coldata$day[22:45] <- "D7"
countdata <- countdata[, rownames(coldata)]
coldata$condition <- as.factor(coldata$condition)

# Build DESeq2 dataset
dds <- DESeqDataSetFromMatrix(countData = countdata,
                              colData   = coldata,
                              design    = ~ condition)
# Filter genes with row sum less than 1
dds <- dds[rowSums(counts(dds)) > 1, ]
barplot(colSums(assay(dds)), las = 3, cex.names = 0.5, col = rep(rainbow(4), each = 3))

# rlog normalization
rld <- rlog(dds, blind = FALSE)
```


# Differential Expression Analysis (DESeq2)
```{r DESeq2_analysis}
# Day 2: condition 3 vs 1 (in vitro PMC vs non-PMC, Day 2)
dds$condition <- relevel(dds$condition, ref = "1")
dds <- DESeq(dds)
res0.05 <- list()
res <- results(dds, name = "condition_3_vs_1")
res0.05[["condition_3_vs_1"]] <- as.data.frame(na.omit(res))

# Day 7: condition 6 vs 4 (in vitro PMC vs non-PMC, Day 7)
dds$condition <- relevel(dds$condition, ref = "4")
dds <- DESeq(dds)
res <- results(dds, name = "condition_6_vs_4")
res0.05[["condition_6_vs_4"]] <- as.data.frame(na.omit(res))
```


# Merge DESeq2 Results Across All Groups
```{r merge_DESeq2_results}
data <- lapply(names(res0.05), function(cluster_name) {
  df <- as.data.frame(res0.05[[cluster_name]])
  df$ID <- rownames(df)
  df <- df[!grepl("^ENSG|^DNAH", df$ID), ]
  df <- df[!grepl("^MT-|\\.|^MT", df$ID), ]
  df <- df[!grepl("^(RPS|RPL)", df$ID), ]
  df <- df[!df$ID %in% housekeeping_genes$Symbol, ]
  df <- df[, c(2, 6, 7), drop = FALSE]
  df$cluster <- cluster_name
  return(df)
})
data <- do.call(rbind, data)
```


# Figure 2F: GSEA Bubble Plot (Bulk RNA-seq, Day 2)
```{r Figure_2F_GSEA}
# Filter significantly upregulated genes
log2FC_cutoff <- log2(1)
padj_cutoff   <- 0.05

condition_3_vs_1 <- data %>% filter(cluster == "condition_3_vs_1") %>%
  filter(log2FoldChange > log2FC_cutoff & padj < padj_cutoff) %>%
  arrange(padj, desc(abs(log2FoldChange))) %>% dplyr::select(ID)

condition_6_vs_4 <- data %>% filter(cluster == "condition_6_vs_4") %>%
  filter(log2FoldChange > log2FC_cutoff & padj < padj_cutoff) %>%
  arrange(padj, desc(abs(log2FoldChange))) %>% dplyr::select(ID)

# Pathway enrichment analysis
geneset <- read.gmt("Reactome_human.gmt")
geneset <- geneset[geneset$term %in% top_markers$geneset, ]

# Load and prepare gene-level data for Day 2
group1 <- data %>% filter(cluster == "condition_3_vs_1")
group1$external_gene_name <- group1$ID
group1$logFC <- group1$log2FoldChange
group1$FDR   <- group1$padj
group1 <- group1[, c("external_gene_name", "logFC", "FDR")]
group1 <- group1[group1$external_gene_name != "", ]
rownames(group1) <- group1$external_gene_name

# Add up/down regulation column
group1$g1 <- "normal"
group1$g1[group1$logFC >  log2(1) & group1$FDR < 0.05] <- "up"
group1$g1[group1$logFC < -log2(1) & group1$FDR < 0.05] <- "down"
table(group1$g1)

# Get differentially expressed genes and sorted gene list
DEGs <- group1[group1$g1 != "normal", ]
DEGs <- DEGs[order(DEGs$logFC, decreasing = TRUE), ]
genelist <- DEGs$logFC
names(genelist) <- DEGs$external_gene_name

# Run GSEA and output all results
egmt <- GSEA(genelist, TERM2GENE = geneset, pvalueCutoff = 0.1,
             pAdjustMethod = "fdr", minGSSize = 1, maxGSSize = 500)
dotplot(egmt, showCategory = 10, label_format = 100, color = "pvalue")

# Prepare enrichment data frame
dt <- data.frame(egmt)
dt$Description <- factor(dt$ID, levels = rev(dt$ID))
dt <- dt %>% filter(NES > 0)

# Basic enrichment bar plot
Fig2_B_sup <- ggplot() +
  geom_bar(data = dt,
           aes(x = NES, y = reorder(Description, NES)),
           width = 0.5,
           fill  = my_color_PMC[4],
           stat  = "identity") +
  theme_classic() +
  scale_x_continuous(expand = c(0, 0)) +
  theme(axis.text.y = element_blank()) +
  geom_text(data = dt,
            aes(x = 0.01, y = Description, label = Description),
            size = 4.5, hjust = 0) +
  geom_text(data = dt,
            aes(x = 0.01, y = Description, label = core_enrichment),
            color     = "darkgrey",
            size      = 4,
            fontface  = "italic",
            hjust     = 0,
            vjust     = 2.3) +
  labs(x = "Normalized Enrichment Score (NES)", y = "Pathways")
Fig2_B_sup

# Sort pathway names as factor for Y-axis
dt <- dt[order(dt$NES, decreasing = TRUE), ]
dt$ID   <- factor(dt$ID, levels = dt$ID)
dt$xlab <- 1:nrow(dt)
summary(dt$NES)

# Pathway names to label in the plot
label      <- dt$ID
data_label <- dt[dt$ID %in% label, ]
data_label$col <- "black"
dt$label_combined <- paste0(dt$ID, ": ", dt$core_enrichment)

p2_F <- ggplot(data = dt, aes(x = xlab, y = NES)) +
  geom_point(aes(size = setSize, fill = -log10(p.adjust)),
             shape = 21, stroke = 0.7, colour = "black") +
  scale_fill_gradient(low = my_color_PMC[4], high = "red") +
  scale_size_continuous(range = c(1, 10)) +
  xlab("") +
  ylab("Normalized Enrichment Score (NES)") +
  theme_classic(base_size = 16) +
  scale_x_continuous(breaks = seq(0, 50, by = 5), labels = seq(0, 50, by = 5)) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.05))) +
  theme(
    legend.text       = element_text(size = 16),
    axis.line         = element_line(color = "black", size = 0.8),
    axis.ticks        = element_line(color = "black", size = 0.8),
    axis.ticks.length = unit(0.2, "cm"),
    axis.text         = element_text(face = "bold")
  ) +
  guides(
    size  = guide_legend(title = "Size"),
    alpha = guide_legend(title = expression("-Log"[10] * " (p-value)"))
  ) +
  geom_text_repel(
    data             = dt,
    aes(x = xlab, y = NES, label = ID),
    size             = 6,
    color            = data_label$col,
    force            = 10,
    point.padding    = 0.5,
    min.segment.length = 0,
    hjust            = 1,
    segment.color    = "grey20",
    segment.size     = 0.1,
    segment.alpha    = 0.8,
    nudge_y          = -0.1
  )
p2_F
```


# Figure 2B: Venn Diagram (scRNA-seq vs Bulk RNA-seq) and Figure 2C: Dot Plot
```{r Figure_2B_2C}
# Filter upregulated bulk RNA-seq DEGs
log2FC_cutoff <- log2(1.5)
padj_cutoff   <- 0.05

condition_3_vs_1 <- data %>% filter(cluster == "condition_3_vs_1") %>%
  filter(log2FoldChange > log2FC_cutoff & padj < padj_cutoff) %>% rownames()

condition_6_vs_4 <- data %>% filter(cluster == "condition_6_vs_4") %>%
  filter(log2FoldChange > log2FC_cutoff & padj < padj_cutoff) %>% rownames()

cluster_marker <- read.csv("Figure 2A PMC vs nonPMC DE genes.csv", header = TRUE)
cluster_marker <- cluster_marker[!grepl("^MT-|^MT|\\.|^(RPS|RPL)", cluster_marker$X), ]
cluster_marker <- cluster_marker[!cluster_marker$X %in% housekeeping_genes$Symbol, ]
cluster_marker <- cluster_marker %>% filter(avg_log2FC > log2(1.5) & p_val_adj < 0.05)

# Shared genes
gene_df <- data.frame(Gene = intersect(condition_3_vs_1, cluster_marker$X))

# Venn diagram
library(Vennerable)
mylist    <- list(PMC = cluster_marker$X, "in vitro PMC" = condition_3_vs_1)
Vennplot  <- Vennerable::Venn(mylist)
w  <- compute.Venn(Vennplot)
gp <- VennThemes(w)
gp[["Face"]][["11"]]$fill    <- "white"
gp[["Face"]][["01"]]$fill    <- "white"
gp[["Face"]][["10"]]$fill    <- "white"
gp[["Set"]][["Set1"]]$col     <- my_color_PMC["PMC"]
gp[["SetText"]][["Set1"]]$col <- my_color_PMC["PMC"]
gp[["Set"]][["Set2"]]$col     <- my_color_PMC["in vitro PMC"]
gp[["SetText"]][["Set2"]]$col <- my_color_PMC["in vitro PMC"]
plot(w, gp = gp)

# Shared genes
intersect(mylist$PMC, mylist$`in vitro PMC`)

# Dot plot of shared genes across clusters
Idents(obj) <- "cluster_num"
my_levels   <- c("CD14", "CD14-active", "CD16", "Intermediate",
                  "Intermediate-metabolic", "Intermediate-interferon", "PMC")
Idents(obj) <- factor(x = Idents(obj), levels = rev(my_levels))

features <- data %>%
  filter(cluster == "condition_3_vs_1") %>%
  filter(ID %in% gene_df$Gene) %>%
  arrange(desc(log2FoldChange), padj) %>%
  dplyr::select(ID)

DefaultAssay(obj) <- "SCT"
p2_C <- DotPlot(obj, features = features$ID, dot.scale = 5) +
  scale_size_area(max_size = 10) +
  RotatedAxis() + xlab("") + ylab("") +
  coord_flip() +
  scale_color_gradientn(colors = colorRampPalette(c("#1F78B4", "white", "#FFED99", "#FF7600"))(10)) +
  scale_y_discrete(guide = guide_axis(angle = -45)) +
  theme(
    axis.text         = element_text(size = 16),
    legend.text       = element_text(size = 16),
    axis.line         = element_line(color = "black", size = 0.8),
    axis.ticks        = element_line(color = "black", size = 0.8),
    axis.ticks.length = unit(0.2, "cm")
  )
p2_C
```


# Figure 2D: Volcano Plot (Bulk RNA-seq, Day 2)
```{r Figure_2D_volcano}
log2FC_cutoff <- log2(1)
padj_cutoff   <- 0.05

need_DEG <- data %>% filter(cluster == "condition_3_vs_1")
need_DEG$significance <- as.factor(
  ifelse(
    need_DEG$padj < padj_cutoff & abs(need_DEG$log2FoldChange) > log2FC_cutoff,
    ifelse(need_DEG$log2FoldChange > log2FC_cutoff, "UP", "DOWN"),
    "NOT"
  )
)

p2_D <- ggplot(data = need_DEG,
               aes(x = log2FoldChange, y = -log10(padj), color = significance)) +
  geom_point(alpha = 0.5, size = 1) +
  theme_classic(base_size = 16) +
  labs(color = "",
       x = expression("log"[2] * " (Fold Change)"),
       y = expression("-log"[10] * " (p-value)")) +
  scale_colour_manual(values = c("#00AFBB", "grey", "#bb0c00")) +
  geom_vline(xintercept = c(-log2FC_cutoff, log2FC_cutoff), lty = 4, col = "grey", lwd = 0.8) +
  geom_hline(yintercept = -log10(padj_cutoff), lty = 4, col = "grey", lwd = 0.8) +
  theme(
    plot.margin       = unit(c(2, 2, 2, 2), "lines"),
    legend.title      = element_blank(),
    legend.position   = "right",
    axis.text         = element_text(size = 16),
    legend.text       = element_text(size = 16),
    axis.line         = element_line(color = "black", size = 0.8),
    axis.ticks        = element_line(color = "black", size = 0.8),
    axis.ticks.length = unit(0.2, "cm")
  ) + xlim(-10, 10)

# Label top 30 DEGs plus specific genes of interest
need_DEG <- need_DEG %>%
  mutate(delabel = ifelse(
    significance == "NOT", NA,
    ifelse(
      ID %in% (ID[significance != "NOT"] %>%
                 .[order(padj[significance != "NOT"],
                         -abs(log2FoldChange[significance != "NOT"]))] %>%
                 head(30)),
      ID, NA
    )
  ))

need_DEG[need_DEG$ID == "IL10",  ]$delabel <- "IL10"
need_DEG[need_DEG$ID == "TGFBI", ]$delabel <- "TGFBI"
need_DEG[need_DEG$ID == "CXCL1", ]$delabel <- "CXCL1"
need_DEG[need_DEG$ID == "CCL2",  ]$delabel <- "CCL2"
need_DEG[need_DEG$ID == "CCL20", ]$delabel <- "CCL20"
need_DEG[need_DEG$ID == "CXCL5", ]$delabel <- "CXCL5"

p2_D <- p2_D + geom_label_repel(
  data = need_DEG,
  aes(log2FoldChange, -log10(padj), label = delabel),
  max.overlaps = Inf,
  size = 4
)
p2_D
```


# Combine and Save Figure 2C–F
```{r Figure_2C-F_patchwork}
library(patchwork)
combined_plot <- p2_D + p2_E + plot_layout(widths = c(2, 1), guides = "collect")
combined_plot <- combined_plot / p2_F + plot_layout(heights = c(1, 1))
combined_plot <- p2_C + combined_plot + plot_layout(widths = c(1, 1))
ggsave(filename = "Figure 2 C-F.pdf", width = 20, height = 13, bg = "white")
```


# Supplementary Figure 2A: GSEA Bar Plot (Day 7 Pathways)
```{r Sup_Figure_2A_GSEA_barplot}
library(ggpubr)

# Volcano plot of gene sets (scPS) for PMC cluster
library(ggprism)
cluster_marker <- read.csv("Sup Fig 1B Cluster-specific gene sets.csv")
cluster_marker <- cluster_marker %>%
  filter(cluster == "PMC" & p_val_adj < 0.05) %>%
  mutate(DEG = case_when(
    avg_log2FC >  log2(1.5) & p_val_adj < 0.05 ~ "up",
    avg_log2FC < -log2(1.5) & p_val_adj < 0.05 ~ "down",
    TRUE ~ "no"
  ))
cluster_marker$label <- ifelse(
  cluster_marker$p_val_adj < log2(3) & abs(cluster_marker$avg_log2FC) > log2(3),
  cluster_marker$gene, ""
)

ggplot(cluster_marker, aes(x = avg_log2FC, y = -log10(p_val_adj), colour = DEG)) +
  geom_point(alpha = 0.85, size = 1.5) +
  scale_color_manual(values = c("steelblue", "gray", "brown")) +
  xlim(c(-5, 5)) +
  geom_vline(xintercept = c(-0.58, 0.58), lty = 4, col = "black", lwd = 0.8) +
  geom_hline(yintercept = -log10(0.05), lty = 4, col = "black", lwd = 0.8) +
  labs(x = "logFC", y = "-log10(FDR)") +
  ggtitle("PMC") +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "right",
        legend.title = element_blank()) +
  geom_label_repel(data = cluster_marker, aes(label = label),
                   size = 3, box.padding = unit(0.5, "lines"),
                   point.padding = unit(0.8, "lines"),
                   segment.color = "black", show.legend = FALSE,
                   max.overlaps = 10000) +
  theme_prism(border = TRUE)

# Bar plot: top PMC pathways by category
merged_data <- merge(
  x    = cluster_marker,
  y    = top_markers,
  by.x = "gene",
  by.y = "geneset",
  all  = FALSE
)
merged_data <- merged_data %>%
  filter(category %in% c("Immune System", "Hemostasis", "Metabolic")) %>%
  filter(gene %in% top_markers$geneset) %>%
  arrange(desc(avg_log2FC), p_val_adj) %>%
  slice_head(n = 30) %>%
  ungroup()

Fig2_A_sup <- ggbarplot(
  merged_data, x = "gene", y = "avg_log2FC",
  fill        = "category",
  color       = "white",
  orientation = "horiz",
  palette     = "nejm",
  legend      = "right",
  sort.val    = "asc",
  sort.by.groups = TRUE
) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_discrete(expand = c(0, 0)) +
  xlab("") +
  ylab(expression("Log"[2] * " (FC)")) +
  ylim(-5, 5.3)
Fig2_A_sup
```


# Supplementary Figure 2B–C: Macrophage Gene Set Overlap and Donut Chart
```{r Sup_Figure_2B_2C_macrophage}
library(VennDiagram)
library(UpSetR)

M_GS <- read_excel("Macrophage gene set.xlsx")
M_GS <- split(M_GS$Marker, M_GS$Macrophages)
M_GS <- lapply(M_GS, unique)

mylist <- list(
  "in vitro PMC"    = intersect(condition_3_vs_1$ID, unlist(M_GS)),
  "in vitro PMC Mφ" = intersect(condition_6_vs_4$ID, unlist(M_GS)),
  M1 = unique(M_GS$M1),
  M2 = unique(M_GS$M2)
)

# UpSet plot
Venn_data <- fromList(mylist)
upset(Venn_data,
  sets        = names(Venn_data),
  keep.order  = TRUE,
  order.by    = "freq",
  point.size  = 3,
  line.size   = 1,
  number.angles = 0,
  text.scale  = 2,
  sets.x.label = "Size"
)

# Define comparison sets for donut chart
sets <- list(
  "in vitro PMC/M1"    = list(A = mylist[["in vitro PMC"]],    B = mylist[["M1"]]),
  "in vitro PMC/M2"    = list(A = mylist[["in vitro PMC"]],    B = mylist[["M2"]]),
  "in vitro PMC Mφ/M1" = list(A = mylist[["in vitro PMC Mφ"]], B = mylist[["M1"]]),
  "in vitro PMC Mφ/M2" = list(A = mylist[["in vitro PMC Mφ"]], B = mylist[["M2"]])
)

# Compute overlap ratios and organize into a data frame
df <- do.call(rbind, lapply(names(sets), function(name) {
  A       <- sets[[name]]$A
  B       <- sets[[name]]$B
  overlap <- length(intersect(A, B))
  total   <- length(A)
  ratio   <- overlap / total
  data.frame(Comparison = name, Overlap = overlap, Total = total, Ratio = ratio)
}))

# Color definitions
my_colors <- c(
  "in vitro PMC/M1"    = "#81D8D0",
  "in vitro PMC/M2"    = "#1F77B4",
  "in vitro PMC Mφ/M1" = "#ECA9B0",
  "in vitro PMC Mφ/M2" = "#D62728"
)

# Compute ymin and ymax for donut chart
df$ymax  <- cumsum(df$Ratio)
df$ymin  <- c(0, head(df$ymax, n = -1))
df$label <- paste0(df$Comparison, "\n(", round(df$Ratio * 100, 1), "%)")

Fig2_C_sup <- ggplot(df, aes(ymax = ymax, ymin = ymin, xmax = 4, xmin = 3, fill = Comparison)) +
  geom_rect(color = "white", size = 1) +
  coord_polar(theta = "y") +
  xlim(c(2, 4)) +
  theme_void() +
  scale_fill_manual(values = my_colors) +
  geom_text(aes(x = 3.5, y = (ymin + ymax) / 2, label = label), size = 5) +
  theme(legend.position = "none")
Fig2_C_sup
```


# Combine and Save Supplementary Figure 2A–C
```{r Sup_Figure_2A-C_patchwork}
library(patchwork)
combined_plot <- (Fig2_B_sup / Fig2_C_sup) + plot_layout(heights = c(1, 2))
Fig2_A_sup + combined_plot
ggsave(filename = "Figure 2 Sup A-C.pdf", width = 10, height = 10, bg = "white")
```


# Supplementary: GSEA of PMC vs Non-PMC (scRNA-seq)
```{r Sup_PMC_GSEA_scRNAseq}
cluster_marker <- read.csv("Figure 2A PMC vs nonPMC DE genes.csv", header = TRUE)
cluster_marker <- cluster_marker[!grepl("^MT-|^MT|\\.|^(RPS|RPL)", cluster_marker$X), ]
cluster_marker <- cluster_marker %>%
  filter(abs(avg_log2FC) > log2(1) & p_val_adj < 0.05) %>%
  filter(pct.1 > 0.1 | pct.2 > 0.1) %>%
  arrange(desc(avg_log2FC), p_val_adj)

# Extract FDR and FC values
group1 <- cluster_marker
group1$external_gene_name <- group1$X
group1$logFC <- group1$avg_log2FC
group1$FDR   <- group1$p_val_adj
group1 <- group1[, c("external_gene_name", "logFC", "FDR")]
group1 <- group1[group1$external_gene_name != "", ]
rownames(group1) <- group1$external_gene_name

# Add up/down regulation column
group1$g1 <- "normal"
group1$g1[group1$logFC >  log2(1) & group1$FDR < 0.05] <- "up"
group1$g1[group1$logFC < -log2(1) & group1$FDR < 0.05] <- "down"
table(group1$g1)

# Get differentially expressed genes and sorted gene list
DEGs <- group1[group1$g1 != "normal", ]
DEGs <- DEGs[order(DEGs$logFC, decreasing = TRUE), ]
genelist <- DEGs$logFC
names(genelist) <- DEGs$external_gene_name

# Load Reactome gene sets
geneset <- read.gmt("Reactome_human.gmt")
geneset <- geneset[geneset$term %in% top_markers$geneset, ]

# Run GSEA and output all results
egmt <- GSEA(genelist, TERM2GENE = geneset, pvalueCutoff = 0.1,
             pAdjustMethod = "fdr", minGSSize = 1, maxGSSize = 500000)
cnetplot(egmt, showCategory = 10, foldChange = genelist,
         circular = TRUE, colorEdge = TRUE, size_item = 5) +
  scale_color_gradientn(colours = c("white", "red"))

dotplot(egmt, showCategory = 10, label_format = 100, color = "pvalue")

# Enrichment bar plot
dt <- data.frame(egmt)
dt$Description <- factor(dt$ID, levels = rev(dt$ID))
dt <- dt %>% filter(NES > 0)

ggplot() +
  geom_bar(data = dt,
           aes(x = NES, y = reorder(Description, NES)),
           width = 0.5, fill = my_color_PMC[4], stat = "identity") +
  theme_classic() +
  scale_x_continuous(expand = c(0, 0)) +
  theme(axis.text.y = element_blank()) +
  geom_text(data = dt,
            aes(x = 0.01, y = Description, label = Description),
            size = 4.5, hjust = 0) +
  geom_text(data = dt,
            aes(x = 0.01, y = Description, label = core_enrichment),
            color = "grey", size = 4, fontface = "italic", hjust = 0, vjust = 2.3) +
  labs(x = "Normalized Enrichment Score (NES)", y = "Pathways")

# Sort pathway names as factor for Y-axis
dt <- dt[order(dt$NES, decreasing = TRUE), ]
dt$ID   <- factor(dt$ID, levels = dt$ID)
dt$xlab <- 1:nrow(dt)
summary(dt$NES)

# Pathway names to label in the plot
label      <- dt$ID
data_label <- dt[dt$ID %in% label, ]
data_label$col <- "black"
dt$label_combined <- paste0(dt$ID, ": ", dt$core_enrichment)

ggplot(data = dt, aes(x = xlab, y = NES)) +
  geom_point(aes(size = setSize, alpha = -log10(p.adjust)),
             shape = 21, stroke = 0.7, fill = my_color_PMC[3], colour = "black") +
  scale_size_continuous(range = c(1, 10)) +
  xlab("") +
  ylab("Normalized Enrichment Score (NES)") +
  theme_classic(base_size = 16) +
  scale_x_continuous(breaks = seq(0, 50, by = 5), labels = seq(0, 50, by = 5)) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.05))) +
  theme(
    legend.text       = element_text(size = 16),
    axis.line         = element_line(color = "black", size = 0.8),
    axis.ticks        = element_line(color = "black", size = 0.8),
    axis.ticks.length = unit(0.2, "cm"),
    axis.text         = element_text(face = "bold")
  ) +
  guides(
    size  = guide_legend(title = "Size"),
    alpha = guide_legend(title = expression("-Log"[10] * " (p-value)"))
  ) +
  geom_text_repel(
    data             = dt,
    aes(x = xlab, y = NES, label = ID),
    size             = 6,
    color            = data_label$col,
    force            = 10,
    point.padding    = 0.5,
    min.segment.length = 0,
    hjust            = 1,
    segment.color    = "grey20",
    segment.size     = 0.1,
    segment.alpha    = 0.8,
    nudge_y          = -0.1
  )
```
