---
title: "Figure 5 and Supplementary Figure 6: Proteomics, NicheNet, and MetaNet"
output: html_document
date: "2024-08-29"
editor_options: 
  chunk_output_type: console
---


# Load Libraries and Define Colors
```{r setup, message=FALSE, warning=FALSE}
library(DEP)
library(factoextra)
library(rafalib)
library(circlize)
library(dendextend)
library(dendsort)
library(ComplexHeatmap)
library(gridBase)
library(pheatmap)
library(psych)
library(clusterProfiler)
library(enrichplot)
library(ggVennDiagram)
library(ggrepel)
library(patchwork)
library(dittoSeq)
library(viridis)
library(scuttle)
library(SingleCellExperiment)
library(nichenetr)
library(OmnipathR)
library(MetaNet)
library(igraph)
library(pcutils)
library(dplyr)
library(tidyverse)
library(ggplot2)

my_color_group <- c(
  "HIV+AS+" = "#ff6b6d",
  "HIV-AS+" = "#ffa725",
  "HIV+AS-" = "#A6CEE3",
  "HIV-AS-" = "lightgrey"
)
```


# Supplementary Figure 6A: Proteomics QC, Normalization, and PCA
```{r Sup_Figure_6A_PCA}
# Load and prepare mass spectrometry data
data        <- read.csv("Figure 5 MassSpec normlized data.csv", header = TRUE, stringsAsFactors = FALSE)
data_unique <- make_unique(data, "Genes", "Protein.Name", delim = ";")

LFQ_columns <- 4:35
meta <- data.frame(Group = rep(c("HIV-AS-", "HIV+AS-", "HIV-AS+", "HIV+AS+"), each = 8))
rownames(meta) <- colnames(data)[LFQ_columns]

experimental_design <- data.frame(
  label     = colnames(data)[LFQ_columns],
  condition = substr(colnames(data)[LFQ_columns], 1, 4),
  replicate = 1:8
)

data_se <- make_se(data_unique, LFQ_columns, experimental_design)

# Filter for proteins identified in all replicates of at least one condition
data_filt <- filter_missval(data_se, thr = 1)

# Bar plot: number of proteins identified per sample
plot_numbers(data_filt)

# VSN normalization and comparison boxplot
data_norm <- normalize_vsn(data_filt)
plot_normalization(data_filt, data_norm)

# PCA
norm_df <- t(data.frame(data_norm@assays@data@listData))
rownames(norm_df) <- rownames(meta)
res.pca <- prcomp(norm_df, center = TRUE, scale. = TRUE)
fviz_eig(res.pca, addlabels = TRUE)

pca_result <- data.frame(
  PC1   = res.pca$x[, 1],
  PC2   = res.pca$x[, 2],
  Group = meta$Group
)

Fig5_Sup_A <- ggplot(data = pca_result, aes(x = PC1, y = PC2, col = Group)) +
  geom_point(size = 3) +
  theme_bw(base_size = 16) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line        = element_blank(),
    axis.ticks       = element_blank(),
    legend.position  = "right"
  ) +
  scale_color_manual(values = my_color_group) +
  labs(x = "PC1 (41%)", y = "PC2 (7.9%)")
Fig5_Sup_A
```


# Figure 5A: Circular Heatmap of log2 Fold-Change Across Groups
```{r Figure_5A_circular_heatmap}
# Test each group versus the ASNN control
data_diff <- test_diff(data_norm, type = "manual",
                       test = c("ASPP_vs_ASNN", "ASNP_vs_ASNN", "ASPN_vs_ASNN"))

# Denote significant proteins: alpha = 0.5, lfc = log2(2)
dep <- add_rejections(data_diff, alpha = 0.5, lfc = log2(2))
data_results <- get_results(dep)

# Extract log2 fold-change matrix
plot_heatmap <- plot_heatmap(dep, type = "contrast",
                             kmeans = FALSE, clustering_distance = "pearson",
                             row_font_size = 10, k = 0, col_limit = 4,
                             show_row_names = TRUE)
dat_matrix           <- plot_heatmap@ht_list$`log2 Fold change`@matrix
colnames(dat_matrix) <- c("HIV+AS-", "HIV-AS+", "HIV+AS+")

# Scale rows for circular heatmap display
cir1   <- t(scale(t(dat_matrix)))
mycol  <- colorRamp2(c(-1.45, 0, 2.27), c("#0da9ce", "white", "#e74a32"))

pdf(file = "Figure 5 A.pdf", width = 6, height = 6)

# Circular layout: gap between first and last sector
circos.par(gap.after = 30)
circos.heatmap(cir1,
  col             = mycol,
  dend.side       = "inside",    # Dendrogram shown on inner ring
  rownames.side   = "outside",   # Row names shown on outer ring
  track.height    = 0.28,
  rownames.col    = "black",
  rownames.cex    = 0.5,
  rownames.font   = 1,
  bg.border       = "black",
  cluster         = TRUE,
  dend.track.height = 0.18,
  dend.callback   = function(dend, m, si) {
    color_branches(dend, k = 3, col = 1:15)  # Color dendrogram branches
  }
)

# Add color legend
lg <- Legend(title     = "Exp",
             col_fun   = mycol,
             direction = "vertical")
grid.draw(lg)

# Add column name labels on the outer track
circos.track(
  track.index = get.current.track.index(),
  panel.fun   = function(x, y) {
    if (CELL_META$sector.numeric.index == 1) {
      cn <- colnames(cir1)
      n  <- length(cn)
      circos.text(
        rep(CELL_META$cell.xlim[2], n) + convert_x(0.5, "mm"),
        1:n + 5,
        cn, cex = 0.6, adj = c(0, 0.5), facing = "inside"
      )
    }
  },
  bg.border = NA
)
circos.clear()
dev.off()
```


# Figure 5B: Pseudobulk Proteomics vs scPS Pathway Score Correlation Heatmap
```{r Figure_5B_correlation_heatmap}
# Re-run with relaxed thresholds to get broader protein list
dep <- add_rejections(data_diff, alpha = 0.5, lfc = log2(1))
data_results <- get_results(dep)
plot_heatmap <- plot_heatmap(dep, type = "contrast",
                             kmeans = FALSE, clustering_distance = "pearson",
                             row_font_size = 10, k = 0, col_limit = 4,
                             show_row_names = TRUE)

# Helper: extract upregulated proteins per group from a ComplexHeatmap object
get_data_select_protein <- function(plot_heatmap,
                                    data_results,
                                    fdr_thresh   = 0.5,
                                    fc_threshold = 1,
                                    slot_name    = "log2 Fold change") {
  dat_matrix <- plot_heatmap@ht_list[[slot_name]]@matrix
  groups     <- colnames(dat_matrix)
  out_list   <- vector("list", length(groups))
  names(out_list) <- groups

  # Build expected FDR column name robustly
  safe_col <- function(g) {
    base <- gsub("[^A-Za-z0-9]+", "_", g)
    paste0(base, "_p.val")
  }

  for (g in groups) {
    group_df <- data.frame(
      external_gene_name = rownames(dat_matrix),
      logFC              = dat_matrix[, g],
      stringsAsFactors   = FALSE
    )

    # Locate FDR column with fallback
    col_try <- safe_col(g)
    if (!col_try %in% colnames(data_results)) {
      alt <- paste0(gsub(" ", "_", g), "_p.val")
      if (alt %in% colnames(data_results)) {
        col_try <- alt
      } else {
        stop(sprintf("FDR column for group '%s' not found. Tried: '%s' and '%s'.",
                     g, safe_col(g), alt))
      }
    }

    fdr_df <- data.frame(
      external_gene_name = data_results$name,
      FDR                = data_results[[col_try]],
      stringsAsFactors   = FALSE
    )

    merged <- merge(group_df, fdr_df, by = "external_gene_name", all.x = TRUE)
    merged <- merged[merged$external_gene_name != "" & !is.na(merged$logFC),
                     c("external_gene_name", "logFC", "FDR")]

    # Label direction
    l2thr        <- log2(fc_threshold)
    merged$g1    <- "normal"
    merged$g1[merged$logFC >  l2thr & merged$FDR < fdr_thresh] <- "up"
    merged$g1[merged$logFC < -l2thr & merged$FDR < fdr_thresh] <- "down"

    dat_up        <- merged[merged$g1 == "up", ]
    dat_up$AS_HIV <- g
    rownames(dat_up) <- dat_up$external_gene_name
    out_list[[g]] <- dat_up
  }

  up_gene_vec <- unique(unlist(lapply(out_list, function(x) x$external_gene_name)))
  return(list(per_group_up = out_list, up_gene_vec = up_gene_vec))
}

# Extract normalized proteomics data for upregulated proteins
MassSpec_data_norm   <- data.frame(data_norm@assays@data@listData)
data_select_protein  <- get_data_select_protein(plot_heatmap, data_results,
                                                fdr_thresh = 0.5, fc_threshold = 1)
MassSpec_data_norm   <- MassSpec_data_norm %>%
  filter(rownames(MassSpec_data_norm) %in% data_select_protein$up_gene_vec)
data1 <- data.frame(t(MassSpec_data_norm))
rownames(data1) <- experimental_design$label

# PMC-specific pathways for scPS pseudobulk scoring
PMC_pathways <- c(
  "Common Pathway of Fibrin Clot Formation",
  "Defects of contact activation system (CAS) and kallikrein/kinin system (KKS)",
  "Diseases of hemostasis",
  "Formation of Fibrin Clot (Clotting Cascade)",
  "Intrinsic Pathway of Fibrin Clot Formation",
  "GP1b-IX-V activation signalling",
  "Chemokine receptors bind chemokines",
  "Prostacyclin signalling through prostacyclin receptor",
  "Inhibition of voltage gated Ca2+ channels via Gbeta/gamma subunits",
  "Gap junction assembly",
  "Transport of connexons to the plasma membrane",
  "Microtubule-dependent trafficking of connexons from Golgi to the plasma membrane",
  "PKA activation in glucagon signalling",
  "Glucagon-type ligand receptors",
  "Glucagon signaling in metabolic regulation",
  "Kinesins"
)

# Extract scPS scores per cell and aggregate per subject × cluster
data_scps <- data.frame(
  t(obj[["scPS"]]@layers$counts[rownames(obj[["scPS"]]) %in% PMC_pathways, ])
)
colnames(data_scps)          <- PMC_pathways
data_scps$HTO_classification <- obj@meta.data$HTO_classification
data_scps$cluster_num        <- obj@meta.data$cluster_num

# Average per subject × cluster
data_scps <- data_scps %>%
  group_by(HTO_classification, cluster_num) %>%
  summarize(across(
    all_of(PMC_pathways),
    ~ mean(.x, na.rm = TRUE),
    .names = "average_{.col}"
  ))
colnames(data_scps) <- gsub("^average_", "", colnames(data_scps))

# Pivot to wide format with cluster_pathway column names
data_cluster_num <- data_scps %>%
  pivot_longer(
    cols     = -c(HTO_classification, cluster_num),
    names_to = "pathway", values_to = "score"
  ) %>%
  mutate(pathway = paste0(cluster_num, "_", pathway)) %>%
  dplyr::select(-cluster_num) %>%
  pivot_wider(names_from = pathway, values_from = score) %>%
  dplyr::select(!starts_with("Intermediate"))

# Pearson correlation between proteomics and scPS pathway scores
data2 <- data_cluster_num[, 2:ncol(data_cluster_num)] %>% as.data.frame()
rownames(data2) <- data_cluster_num$HTO_classification
data2 <- data2[rownames(data1), ]

res <- corr.test(data1, data2, use = "pairwise", method = "pearson",
                 adjust = "fdr", alpha = 0.05)
pmt <- res$p.adj
cmt <- res$r

# Filter proteins with at least 5 significant correlations (FDR < 0.5)
count_below <- apply(pmt, 1, function(row) sum(row < 0.5, na.rm = TRUE))
count_below <- data.frame(protein = names(count_below), count = count_below)
count_below <- count_below %>% filter(count > 5)

count_below %>%
  ggplot(aes(reorder(protein, count), count)) +
  geom_col(aes(fill = count)) +
  coord_flip() +
  labs(x = "Protein") +
  theme_classic() +
  scale_y_continuous(breaks = seq(0, ncol(data_cluster_num) - 2, by = 5)) +
  ggtitle("Proteins with count > 5 (adjusted p-value < 0.5)")

# Subset correlation and p-value matrices to filtered proteins
cmt.1 <- cmt[count_below$protein, ]
pmt.1 <- pmt[count_below$protein, ]

# Convert p-values to significance symbols
if (!is.null(pmt.1)) {
  ssmt        <- pmt.1 < 0.05
  smt         <- pmt.1 > 0.05 & pmt.1 < 0.5
  pmt.1[ssmt] <- "**"
  pmt.1[smt]  <- "*"
  pmt.1[!ssmt & !smt] <- ""
} else {
  pmt.1 <- FALSE
}
pmt.1 <- as.data.frame(pmt.1) %>% dplyr::select(where(~ !any(is.na(.))))
cmt.1 <- cmt.1[, colnames(pmt.1)]

# Annotation data frame for cluster color strip
dfh            <- data.frame(group = sub("_.*", "", colnames(cmt.1)))
rownames(dfh)  <- colnames(cmt.1)
node_colors    <- setNames(
  c("#FDBF6F", "#FF7F00", "#B15928", "#CCAFCA"),
  c("CD14", "CD14-active", "CD16", "PMC")
)

# Save heatmap
pdf(file = "Figure 5 B.pdf", width = 40, height = 18)
pheatmap::pheatmap(
  t(cmt.1),
  scale                    = "none",
  color                    = colorRampPalette(c("navy", "white", "red"))(50),
  annotation_row           = dfh,
  annotation_colors        = list(group = node_colors),
  cluster_rows             = TRUE,
  cutree_rows              = 1,
  cutree_cols              = 2,
  fontsize                 = 20,
  number_color             = "black",
  clustering_distance_row  = "binary",
  display_numbers          = t(pmt.1),
  fontsize_number          = 16,
  angle_col                = 45
)
dev.off()
```


# Figure 5B: Circular Correlation Heatmap (circos)
```{r Figure_5B_circos_heatmap}
plot_df <- as.matrix(cmt.1)

circos.clear()
circos.par(start.degree = -30, gap.degree = 100, track.height = 0.4)
circos.heatmap(plot_df,
  col                = colorRamp2(c(-0.5, 0, 0.5), c("blue", "white", "red")),
  cell.lwd           = 1,
  bg.border          = "black",
  bg.lty             = par("lty"),
  bg.lwd             = par("lwd"),
  cluster            = TRUE,
  clustering.method  = "ward.D2",
  distance.method    = "euclidean",
  dend.side          = "inside",
  dend.track.height  = 0.1,
  rownames.side      = "outside",
  rownames.cex       = 1,      # Row name font size
  rownames.font      = par("font"),
  cell.border        = NA,
  rownames.col       = "black"
)

# Add column name labels on the outer ring
circos.track(
  track.index = get.current.track.index(),
  panel.fun = function(x, y) {
    if (CELL_META$sector.numeric.index == 1) {
      cn <- colnames(plot_df)
      n  <- length(cn)
      circos.text(
        rep(CELL_META$cell.xlim[2], n) + 0.5 * convert_x(1, "mm"),
        0.5 * (1:n) + 1.2,
        cn, adj = c(0, 0.5), cex = 1, facing = "inside"
      )
    }
  },
  bg.border = NA
)

# Add correlation legend (color matches the heatmap)
lgd <- Legend(
  title           = "Correlation",
  title_position  = "leftcenter-rot",
  title_gp        = gpar(fontsize = 10),
  col_fun         = colorRamp2(c(-0.5, 0, 0.5), c("blue", "white", "red")),
  direction       = "vertical",
  border          = "black",
  labels_gp       = gpar(fontsize = 8, col = "black", font = 3),
  legend_width    = unit(0.5, "cm"),
  legend_height   = unit(0.5, "cm")
)
grid.draw(lgd)
```


# Figure 5C: NicheNet Ligand-Receptor Analysis
```{r Figure_5C_NicheNet}
# Load NicheNet pre-computed databases
lr_network          <- readRDS("NicheNet/lr_network_human_21122021.rds")
signal_network      <- readRDS("NicheNet/signaling_network_human_21122021.rds")
ligand_target_matrix <- readRDS("NicheNet/ligand_target_matrix_nsga2r_final.rds")
weighted_networks   <- readRDS("NicheNet/weighted_networks_nsga2r_final.rds")

# Database summary
cat("Number of ligand-receptor pairs:", nrow(lr_network), "\n")
cat("Number of unique ligands:", length(unique(lr_network$from)), "\n")
cat("Number of unique receptors:", length(unique(lr_network$to)), "\n")
cat("Ligand-target matrix dimensions:", dim(ligand_target_matrix), "\n")

# Check whether genes of interest are present in the signaling network
genes_check <- c("VAMP8", "RAB2A", "GAK", "ORAI1", "PTGR1", "ADRM1", "ZDHHC20",
                 "ADCY6", "STIM1", "PAG1", "CD3G", "PTPN11", "TRPC6")
list(
  present_in_from = genes_check[genes_check %in% signal_network$from],
  present_in_to   = genes_check[genes_check %in% signal_network$to],
  missing_genes   = genes_check[!genes_check %in% c(signal_network$from, signal_network$to)]
)

# Build gene set for each signaling hub gene and export as GMT
dat_sig <- signal_network[signal_network$from %in% genes_check, ]
gs      <- lapply(split(dat_sig$to, dat_sig$from), unique)

write_gmt <- function(gs, file) {
  sink(file)
  lapply(names(gs), function(i) {
    cat(paste(c(i, "tmp", gs[[i]]), collapse = "\t"))
    cat("\n")
  })
  sink()
}
write_gmt(gs, "NicheNet/NicheNet.gmt")

# Dot plot of NicheNet gene set scores per cluster × group
DefaultAssay(obj_small) <- "NicheNet"
obj_small$group <- paste0(obj_small$cluster_num, obj_small$AS_HIV, "_")
Idents(obj_small) <- "group"

DotPlot(obj_small, features = names(predBind), dot.scale = 5) +
  scale_size_area(max_size = 10) +
  RotatedAxis() + xlab("") + ylab("") +
  scale_color_gradientn(colors = colorRampPalette(c("#1F78B4", "white", "#FFED99", "#FF7600"))(10)) +
  theme_classic(base_size = 25) +
  theme(axis.title.x = element_blank(),
        axis.text.x  = element_text(angle = 90, hjust = 1))

# Aggregate NicheNet scores across cells per cluster × group
features <- names(predBind)
metadata <- obj_small@meta.data
sce      <- SingleCellExperiment(
  assays  = list(counts = obj_small@assays$NicheNet@data),
  colData = metadata
)

celltype_mean <- aggregateAcrossCells(
  as(sce, "SingleCellExperiment"),
  id            = DataFrame(label = sce$cluster_num, sample = sce$AS_HIV),
  statistics    = "mean",
  use.assay.type = "counts",
  subset.row    = features
)

columnsToUse   <- c("cluster_num", "AS_HIV")
colData(sce)   <- colData(sce) %>% data.frame() %>%
  dplyr::select(all_of(columnsToUse)) %>% DataFrame()
colnames(celltype_mean) <- paste0("cell", seq_len(ncol(celltype_mean)))

# NicheNet heatmap
p_pheatmap <- dittoHeatmap(
  celltype_mean,
  assay                  = "counts",
  cluster_cols           = TRUE,
  cluster_row            = TRUE,
  scaled.to.max          = FALSE,
  border_color           = "grey",
  complex                = FALSE,
  clustering_distance_rows = "euclidean",
  clustering_method      = "complete",
  cutree_cols            = 3,
  scale                  = "row",
  heatmap.colors         = colorRampPalette(c("lightblue", "white", "red"))(25),
  annot.by               = c("cluster_num", "AS_HIV"),
  annotation_colors      = list(
    AS_HIV      = my_color_group,
    cluster_num = my_color[c(4, 5, 7, 8)],
    ncells      = plasma(100)
  )
)
p_pheatmap

pdf("Figure 5C NicheNet.pdf", width = 10, height = 4)
p_pheatmap
dev.off()
```


# Supplementary Figure 6B: Venn Diagram of Upregulated Proteins Across Groups
```{r Sup_Figure_6B_Venn}
dat_matrix           <- plot_heatmap@ht_list$`log2 Fold change`@matrix
colnames(dat_matrix) <- c("HIV+AS-", "HIV-AS+", "HIV+AS+")
Groups   <- c("HIV+AS-", "HIV-AS+", "HIV+AS+")

# Proteins upregulated (log2FC > 0) in each group
Genesets <- setNames(
  lapply(Groups, function(group) rownames(dat_matrix)[dat_matrix[, group] > log2(1)]),
  Groups
)

# Shared proteins across all groups
Reduce(intersect, Genesets)
which(sapply(Genesets, function(x) "ADRM1" %in% x))

Fig5_Sup_B <- ggVennDiagram(Genesets,
  category.names      = names(Genesets),
  set_color           = c("#A6CEE3", "#ffa725", "#ff6b6d"),
  set_size            = 3,
  label               = "count",
  label_color         = "black",
  label_alpha         = 0,
  label_size          = 5,
  label_percent_digit = 1,
  edge_size           = 1
) + scale_fill_gradient(low = "white", high = "white")
Fig5_Sup_B
ggsave(filename = "Sup Figure 6 B.pdf", width = 5, height = 5, bg = "white")
```


# Supplementary Figure 6C–H: Per-Group GSEA Bar Plots and Network Plots
```{r Sup_Figure_6C-H_GSEA}
geneset <- read.gmt("Reactome_human.gmt")
geneset <- geneset[geneset$term %in% top_markers$geneset, ]
Groups  <- c("HIV+AS-", "HIV-AS+", "HIV+AS+")

# Helper: run GSEA and return barplot + cnetplot for one group
run_gsea_plot <- function(i, dat_matrix, Groups, Genesets, geneset, my_color_group) {
  # Prepare ranked gene list
  genelist <- dat_matrix[Genesets[[Groups[i]]], Groups[i]]
  genelist <- genelist[order(genelist, decreasing = TRUE)]

  # Run GSEA
  egmt <- GSEA(genelist,
               TERM2GENE      = geneset,
               pvalueCutoff   = 1,
               pAdjustMethod  = "fdr",
               minGSSize      = 1,
               maxGSSize      = 500000)

  # Prepare result data frame
  dt             <- data.frame(egmt)
  dt$Description <- dt$ID
  dt$Description <- factor(dt$Description, levels = rev(dt$Description))
  dt             <- dt %>% filter(NES > 0) %>% arrange(desc(NES)) %>% slice_head(n = 10)

  # Bar plot
  p_bar <- ggplot() +
    geom_bar(data = dt,
             aes(x = NES, y = reorder(Description, NES)),
             width = 0.5, fill = my_color_group[Groups[i]], stat = "identity") +
    theme_classic() +
    scale_x_continuous(expand = c(0, 0)) +
    theme(axis.text.y = element_blank()) +
    geom_text(data = dt,
              aes(x = 0.01, y = Description, label = Description),
              size = 4.5, hjust = 0) +
    geom_text(data = dt,
              aes(x = 0.01, y = Description, label = core_enrichment),
              color = "darkgrey", size = 4, fontface = "italic", hjust = 0, vjust = 2.3) +
    labs(x = "Normalized Enrichment Score (NES)", y = "Pathways")

  # cnetplot
  p_net <- cnetplot(egmt,
                    showCategory = 10,
                    foldChange   = genelist,
                    circular     = TRUE,
                    colorEdge    = TRUE,
                    size_item    = 5) +
    scale_color_gradientn(colours = c("white", "red"))

  return(list(bar = p_bar, net = p_net, gsea = dt))
}

# Run for all three groups
results  <- lapply(1:3, run_gsea_plot,
                   dat_matrix    = dat_matrix,
                   Genesets      = Genesets,
                   Groups        = Groups,
                   geneset       = geneset,
                   my_color_group = my_color_group)

p_list   <- lapply(results, function(x) x$bar)
p2_list  <- lapply(results, function(x) x$net)
egmt_res <- lapply(lapply(results, function(x) x$gsea), rownames)

# Shared pathways across all three groups
Reduce(intersect, egmt_res)

# Combine and save
combined_plot_0 <- Fig5_Sup_A + Fig5_Sup_B + plot_layout(widths = c(1, 1))
combined_plot_1 <- p_list[[1]] + p2_list[[1]] + plot_layout(widths = c(1.5, 1))
combined_plot_2 <- p_list[[2]] + p2_list[[2]] + plot_layout(widths = c(1.5, 1))
combined_plot_3 <- p_list[[3]] + p2_list[[3]] + plot_layout(widths = c(1.5, 1))
combined_plot   <- combined_plot_1 / combined_plot_2 / combined_plot_3 +
  plot_layout(guides = "keep")

ggsave("Sup Figure 6.pdf", width = 15, height = 18)
combined_plot <- combined_plot_0 / combined_plot + plot_layout(heights = c(1, 6))
combined_plot
```


# Supplementary Figure 6: Enrichment Bubble Plot (HIV+AS+ Upregulated Proteins)
```{r Sup_Figure_6_enrichment_bubble}
enrichplot_1 <- enricher(
  gene          = data_select_protein$up_gene_vec,
  pvalueCutoff  = 0.5,
  qvalueCutoff  = 0.5,
  TERM2GENE     = geneset
)
dotplot(enrichplot_1, x = "GeneRatio", color = "p.adjust", showCategory = 10)

combined_df             <- data.frame(enrichplot_1)
combined_df$Description <- str_wrap(combined_df$Description, width = 50)

# Bubble plot
ggplot(combined_df, aes(x = GeneRatio, y = fct_reorder(Description, GeneRatio))) +
  geom_point(aes(size = Count, fill = p.adjust), shape = 21, color = "white") +
  scale_fill_gradient(low = "#E27371", high = "#5D82A7") +
  labs(title = "", y = "", x = "GeneRatio") +
  guides(fill = guide_colorbar(reverse = TRUE, order = 1)) +
  theme_bw(base_size = 18) +
  theme(
    axis.text.y      = element_text(color = "black"),
    strip.text       = element_text(hjust = 0.5, color = "black"),
    strip.background = element_rect(fill = "white"),
    axis.line        = element_blank(),
    axis.ticks       = element_blank(),
    axis.text        = element_text(size = 16),
    legend.text      = element_text(size = 16)
  )

ggsave(filename = "MassSpec Reactome Enrichment.pdf", width = 10, height = 13.5, bg = "white")

# Enrichment bar plot
dt             <- data.frame(enrichplot_1)
dt$Description <- factor(dt$Description, levels = rev(dt$Description))
dt$GeneRatio   <- sapply(dt$GeneRatio, function(x) eval(parse(text = x)))

Fig5_B_sup <- ggplot() +
  geom_bar(data = dt,
           aes(x = GeneRatio, y = reorder(Description, GeneRatio)),
           width = 0.5, fill = "orange", stat = "identity") +
  theme_classic() +
  scale_x_continuous(expand = c(0, 0)) +
  theme(axis.text.y = element_blank()) +
  geom_text(data = dt,
            aes(x = 0.01, y = Description, label = Description),
            size = 4.5, hjust = 0) +
  geom_text(data = dt,
            aes(x = 0.01, y = Description, label = geneID),
            color = "darkgrey", size = 4, fontface = "italic", hjust = 0, vjust = 2.3) +
  labs(x = "GeneRatio", y = "Pathways")
Fig5_B_sup

# Sort and label pathways for bubble plot
dt       <- dt[order(dt$GeneRatio, decreasing = TRUE), ]
dt$ID    <- factor(dt$ID, levels = dt$ID)
dt$xlab  <- 1:nrow(dt)
label      <- dt$ID
data_label <- dt[dt$ID %in% label, ]
data_label$col <- "black"

ggplot(data = dt, aes(x = xlab, y = GeneRatio)) +
  geom_point(aes(size = Count, alpha = -log10(p.adjust)),
             shape = 21, stroke = 0.7, fill = my_color_PMC[3], colour = "black") +
  scale_size_continuous(range = c(1, 10)) +
  xlab("") + ylab("GeneRatio") +
  theme_classic(base_size = 16) +
  scale_x_continuous(breaks = seq(0, 50, by = 5), labels = seq(0, 50, by = 5)) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.05))) +
  theme(
    legend.text       = element_text(size = 16),
    axis.line         = element_line(color = "black", size = 0.8),
    axis.ticks        = element_line(color = "black", size = 0.8),
    axis.ticks.length = unit(0.2, "cm"),
    axis.text         = element_text(face = "bold")
  ) +
  guides(
    size  = guide_legend(title = "Size"),
    alpha = guide_legend(title = expression("-Log"[10] * " (p-value)"))
  ) +
  geom_text_repel(data = dt,
                  aes(x = xlab, y = GeneRatio, label = ID),
                  size               = 6,
                  color              = data_label$col,
                  force              = 10,
                  point.padding      = 0.5,
                  min.segment.length = 0,
                  hjust              = 1,
                  segment.color      = "grey20",
                  segment.size       = 0.1,
                  segment.alpha      = 0.8,
                  nudge_y            = -0.1)

cnetplot(enrichplot_1, showCategory = 30,
         circular = TRUE, colorEdge = TRUE, size_item = 5)
```


# Supplementary Figure 5: MetaNet Multi-Omics Correlation Network
```{r Sup_Figure_5_MetaNet}
library(MetaNet)
library(igraph)
library(pcutils)

# Build pseudobulk gene expression matrix per subject × cluster using PMC pathway genes
gene_list <- predBind[names(predBind) %in% PMC_pathways]
all_genes <- unique(unlist(lapply(geneIds(gene_list), identity)))
all_genes <- c(all_genes, intersect(mylist$PMC, mylist$`in vitro PMC`))

# Extract SCT expression data and aggregate per subject × cluster
data_genes <- data.frame(
  t(obj[["SCT"]]@data[rownames(obj[["SCT"]]) %in% all_genes, ])
)
data_genes$HTO_classification <- obj@meta.data$HTO_classification
data_genes$cluster_num        <- obj@meta.data$cluster_num

data_genes <- data_genes %>%
  group_by(HTO_classification, cluster_num) %>%
  summarize(across(
    all_of(colnames(data_genes)[!(colnames(data_genes) %in% c("HTO_classification", "cluster_num"))]),
    ~ mean(.x, na.rm = TRUE),
    .names = "average_{.col}"
  ))
colnames(data_genes) <- gsub("^average_", "", colnames(data_genes))

# Pivot to wide format with cluster_gene column names
data_PMC <- data_genes %>%
  pivot_longer(
    cols     = -c(HTO_classification, cluster_num),
    names_to = "gene", values_to = "score"
  ) %>%
  mutate(gene = paste0(cluster_num, "_", gene)) %>%
  dplyr::select(-cluster_num) %>%
  pivot_wider(names_from = gene, values_from = score) %>%
  as.data.frame()

rownames(data_PMC) <- data_PMC$HTO_classification
data_PMC <- data_PMC[, -1]
data_PMC <- data_PMC[rownames(data1), ]
data_PMC <- data_PMC %>% dplyr::select(!starts_with("Intermediate"))

# Cluster annotation for transcriptomic nodes
data_PMC_g            <- data.frame(cluster = sub("_.*", "", colnames(data_PMC)))
rownames(data_PMC_g)  <- colnames(data_PMC)

# Build multi-omics network: proteomics vs transcriptomics (Spearman, |r| >= 0.6, p < 0.05)
multi1 <- multi_net_build(
  list(Proteomic = data1, Transcriptome = data_PMC),
  method      = "spearman",
  r_threshold = 0.6,
  p_threshold = 0.05
)

# Annotate nodes with cluster class and abundance
multi1_with_anno <- c_net_set(multi1, data_PMC_g, vertex_class = "cluster")
multi1_with_anno <- c_net_set(multi1_with_anno,
                              data.frame("Abundance1" = colSums(data_PMC)),
                              vertex_size = paste0("Abundance", 1))

# Shared plot parameters
node_colors <- setNames(
  c("#b3de69", "#FDBF6F", "#FF7F00", "#B15928", "#CCAFCA"),
  c("Proteomic", "CD14", "CD14-active", "CD16", "PMC")
)
params_list <- list(
  labels_num           = 100,
  vertex.color         = node_colors,
  vertex_size_range    = c(3, 10),
  vertex.label.color   = "black",
  edge_width_range     = c(0.5, 3),
  edge.color           = c("#80b1d3", "#fb8072"),
  edge.curved          = 0.5,
  legend               = TRUE,
  legend_number        = TRUE,
  group_legend_order   = c("Proteomic", "Transcriptome"),
  group_legend_title   = c("Proteomic", "Transcriptome"),
  edge_legend_title    = "Correlation",
  edge_legend_order    = c("positive", "negative"),
  size_legend          = TRUE,
  size_legend_title    = "Abundance",
  width_legend         = TRUE,
  width_legend_title   = "abs(r)",
  lty_legend           = TRUE,
  lty_legend_title     = "Omics relationship",
  gray_color           = "gray80"
)

# Filter to inter-omics edges only and plot
multi2 <- c_net_filter(multi1_with_anno, v_group %in% c("Proteomic", "Transcriptome")) %>%
  c_net_filter(., e_class == "inter", mode = "e")

pdf(file = "Sup Figure 5 MetaNet inter-omics network.pdf", width = 7.5, height = 5, bg = "white")
c_net_plot(multi2, params_list = params_list)
dev.off()

# Export edge table
edges_df <- igraph::as_data_frame(multi2, what = "edges")
write.csv(edges_df, "Sup Figure 6A MetaNet.csv", row.names = FALSE)

# Full network plot (all edges)
pdf(file = "Sup Figure 5 MetaNet full network.pdf", width = 8, height = 5, bg = "white")
c_net_plot(multi1_with_anno, params_list = params_list)
dev.off()

# Highlight 2nd-order neighborhood for selected PMP proteins
PMPs_proteins <- c("ADCY6", "TRPC6", "PTPN11", "VAV1", "ORAI1")

params_highlight <- list(
  labels_num         = 0,
  vertex_size_range  = c(3, 10),
  vertex.label.color = "#d9d9d9",
  edge_width_range   = c(0.5, 3),
  edge.color         = c("#80b1d3", "#fb8072"),
  edge.curved        = 0.5,
  legend             = TRUE,
  group_legend_order = c("Proteomic", "Transcriptome"),
  group_legend_title = c("Proteomic", "Transcriptome"),
  edge_legend_title  = "Correlation",
  edge_legend_order  = c("positive", "negative"),
  lty_legend_title   = "Omics relationship"
)

# Example: highlight neighborhood for first PMP protein
i     <- 1
nodes <- PMPs_proteins[i]
nodes <- V(c_net_neighbors(multi1_with_anno, nodes, order = 2))$name
nodes <- c(PMPs_proteins[i], nodes[grepl("_", nodes)])
g_hl  <- c_net_highlight(multi1_with_anno, nodes = nodes)
c_net_plot(g_hl, params_list = params_highlight, main = "Network")
```
