---
title: "Figure 6: In Vitro PMC vs Non-PMC"
output: html_document
date: "2025-02-05"
editor_options: 
  chunk_output_type: console
---


# Load Libraries
```{r setup, message=FALSE, warning=FALSE}
library(rstatix)
library(dplyr)
library(tidyr)
library(tidyverse)
library(purrr)
library(ggplot2)
library(ggpubr)
library(ggrepel)
library(patchwork)
library(readxl)
library(openxlsx)
library(Cairo)
library(FactoMineR)
library(factoextra)
library(psych)
library(viridis)
library(corrplot)
library(ggsci)
library(clusterProfiler)
library(pheatmap)
library(gridExtra)
library(grid)
```


# Figure 6A–B: Cytokine and Pathway Dot Plots

## Figure 6A: Cytokine Gene Expression (SCT)
```{r Figure_6A_cytokine_dotplot}
p1_1 <- DotPlot(obj, features = c("TNF", "IL1B", "IL6", "IL10", "TGFB1"),
                assay = "SCT", dot.scale = 5) +
  scale_size_area(max_size = 5) +
  coord_flip() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, size = 20, vjust = 1, hjust = 1)) +
  xlab("") + ylab("") +
  guides(size = guide_legend("Percent Expression")) +
  scale_color_gradientn(colors = colorRampPalette(c("#1F78B4", "white", "#FFED99", "#FF7600"))(10)) +
  scale_x_discrete(guide = guide_axis(angle = 0)) +
  theme(
    axis.text        = element_text(size = 20),       # Axis text style
    legend.text      = element_text(size = 16),        # Legend text style
    legend.title     = element_text(size = 16),
    axis.line        = element_line(color = "black", size = 0.8),
    axis.ticks       = element_line(color = "black", size = 0.8),
    axis.ticks.length = unit(0.2, "cm")
  )
p1_1
```

## Figure 6B: IL-10 and TGF-β Pathway Scores (scPS)
```{r Figure_6B_pathway_dotplot}
features <- c(
  "Interleukin-10 signaling",
  rownames(obj[["scPS"]])[grep("IL10", rownames(obj[["scPS"]]))],
  rownames(obj[["scPS"]])[grep("TGF",  rownames(obj[["scPS"]]))]
)
features <- features[features != "FCGR3A-mediated IL10 synthesis"]
features <- features[features != "Downregulation of TGF-beta receptor signaling"]

Idents(obj)       <- "cluster_num"
DefaultAssay(obj) <- "scPS"
Idents(obj)       <- "PMC"

p1 <- DotPlot(obj, features = features, assay = "scPS", dot.scale = 5) +
  scale_size_area(max_size = 5) +
  coord_flip() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.text.y = element_text(angle = 0,  vjust = 1, hjust = 1)) +
  xlab("") + ylab("") +
  guides(size = guide_legend("Percent Expression")) +
  scale_color_gradientn(colors = colorRampPalette(c("#1F78B4", "white", "#FFED99", "#FF7600"))(10)) +
  theme(
    axis.text         = element_text(size = 20),       # Axis text style
    legend.text       = element_text(size = 16),        # Legend text style
    legend.title      = element_text(size = 16),
    axis.line         = element_line(color = "black", size = 0.8),
    axis.ticks        = element_line(color = "black", size = 0.8),
    axis.ticks.length = unit(0.2, "cm")
  )
p1
```


# Figure 6C: PCA Biplot of IL-10 and TGF-β Pathway Genes
```{r Figure_6C_PCA}
# Gather gene universe: IL-10 and TGF-β pathway members
genes <- unique(c(
  geneIds(predBind[["Interleukin-10 signaling"]]),
  geneIds(predBind[["Signaling by TGF-beta Receptor Complex"]]),
  geneIds(predBind[["Signaling by TGFB family members"]]),
  geneIds(predBind[["TGF-beta receptor signaling activates SMADs"]]),
  geneIds(predBind[["TGF-beta receptor signaling in EMT (epithelial to mesenchymal transition)"]])
))
genes <- sort(intersect(genes, rownames(rld)), decreasing = TRUE)

data <- data.frame(t(assay(rld)[genes, c(1:7, 15:29, 38:45)]))
data$group <- rep(
  c("in vitro nonPMC", "in vitro PMC", "in vitro Mφ", "in vitro PMC Mφ"),
  c(7, 7, 8, 8)
)

# PCA
res.pca <- PCA(data[, -ncol(data)], graph = FALSE)
print(res.pca)
get_eigenvalue(res.pca)
fviz_eig(res.pca)

# PCA biplot with confidence ellipses, colored by group
p2 <- fviz_pca_biplot(res.pca,
  geom.ind      = "point",
  pointshape    = 21,
  pointsize     = 2.5,
  labelsize     = 6,
  fill.ind      = data$group,
  col.ind       = data$group,
  select.var    = list(name = c("IL10", "TGFB1", "STAT3", "STAT1",
                                "TNF", "IL6", "IL1B", "CXCL10", "IL22")),
  addEllipses   = TRUE,
  ellipse.type  = "confidence",
  legend.title  = list(fill = "Groups", color = "Groups"),
  repel         = TRUE
) +
  scale_fill_manual(values  = my_color_PMC) +
  scale_color_manual(values = my_color_PMC) +
  theme_cowplot() +
  ggtitle("") +
  xlab("PC1 (37.1%)") + ylab("PC2 (14%)") +
  theme(
    axis.text         = element_text(size = 20),
    axis.title        = element_text(size = 20),
    legend.text       = element_text(size = 20),
    axis.line         = element_line(color = "black", size = 0.8),
    axis.ticks        = element_line(color = "black", size = 0.8),
    axis.ticks.length = unit(0.2, "cm")
  )
p2
# Save: width = 800, height = 600
```


# Figure 6D: TGF-β and IL-10 Gene Expression in Bulk RNA-seq

## Helper: Build Significance Bracket Data Frame
```{r bracket_helper}
# Shared function: extracts Tukey significant comparisons and computes bracket positions
make_bracket_df <- function(tukey_result, group_levels, y_data, bracket_height = 0.002, y_step = 0.01) {
  dat_sig <- as.data.frame(tukey_result$Group) %>%
    mutate(Comparison = rownames(.)) %>%
    rename(p_adj = `p adj`) %>%
    filter(p_adj < 0.05) %>%
    separate(Comparison, into = c("xmin", "xmax"), sep = "-") %>%
    mutate(
      xmin        = trimws(xmin),
      xmax        = trimws(xmax),
      annotations = signif(p_adj, digits = 2)
    ) %>%
    left_join(
      data.frame(Group = names(tapply(y_data, group_levels, max)),
                 y     = tapply(y_data, group_levels, max, na.rm = TRUE)),
      by = c("xmin" = "Group")
    ) %>% rename(ymin = y) %>%
    left_join(
      data.frame(Group = names(tapply(y_data, group_levels, max)),
                 y     = tapply(y_data, group_levels, max, na.rm = TRUE)),
      by = c("xmax" = "Group")
    ) %>%
    rename(ymax = y) %>%
    mutate(y.position = pmax(ymin, ymax))

  dat_sig <- dat_sig %>%
    mutate(
      xstart = as.numeric(factor(xmin, levels = group_levels)),
      xend   = as.numeric(factor(xmax, levels = group_levels)),
      xmid   = (xstart + xend) / 2
    ) %>%
    mutate(y.position = y.position + row_number() * y_step * max(y.position))

  dat_sig %>% mutate(
    y_top  = y.position + bracket_height * y.position,
    y_text = y_top + 0.005 * y.position    # Shift text label slightly upward
  )
}
```

## TGF-β Gene Set Violin Plot
```{r Figure_6D_TGFb}
# Extract TGFb gene rows for selected samples
uni_matrix        <- data.frame(t(assay(rld)[rownames(rld)[grepl("TGFB", rownames(rld))],
                                              c(1:7, 15:29, 38:45)]))
uni_matrix$sum    <- rowSums(uni_matrix)
uni_matrix$median <- apply(uni_matrix, 1, median)

my_sample_col <- data.frame(
  Group = rep(c("in vitro nonPMC", "in vitro PMC", "in vitro Mφ", "in vitro PMC Mφ"), c(7, 7, 8, 8))
)
rownames(my_sample_col) <- rownames(uni_matrix)
uni_matrix <- cbind(my_sample_col, uni_matrix)
uni_matrix$Group <- factor(uni_matrix$Group,
                           levels = c("in vitro nonPMC", "in vitro PMC",
                                      "in vitro Mφ", "in vitro PMC Mφ"))
uni_matrix$Feature <- "TGFb gene set"

# Normality check
model <- lm(sum ~ Group, data = uni_matrix)
ggqqplot(residuals(model))
shapiro_test(residuals(model))

# One-way ANOVA and Tukey HSD
anova <- aov(sum ~ Group, data = uni_matrix)
summary(anova)
tukey <- TukeyHSD(anova)
print(tukey)

# Build bracket data frame for significance annotations
group_levels <- levels(uni_matrix$Group)
bracket_df <- make_bracket_df(tukey, group_levels, uni_matrix$sum,
                              bracket_height = 0.002, y_step = 0.01)

p3 <- ggplot(uni_matrix, aes(x = Group, y = sum, fill = Group)) +
  geom_violin(width = 0.2, alpha = 1, aes(color = Group)) +
  geom_quasirandom(size = 0.1, alpha = 1, color = "grey40") +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 1,
               color = "lightgrey", linewidth = 0.2) +
  scale_fill_manual(values  = my_color_PMC) +
  scale_color_manual(values = my_color_PMC) +
  facet_wrap(~ Feature, scales = "free_y") +
  theme_classic(base_size = 20) +
  labs(x = "", y = "TGFb Gene Set") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text   = element_text(color = "black"),
        axis.title  = element_text(color = "black")) +
  # Left bracket arm
  geom_segment(data = bracket_df,
               aes(x = xstart, xend = xstart, y = y.position, yend = y_top),
               inherit.aes = FALSE) +
  # Right bracket arm
  geom_segment(data = bracket_df,
               aes(x = xend, xend = xend, y = y.position, yend = y_top),
               inherit.aes = FALSE) +
  # Horizontal bracket top
  geom_segment(data = bracket_df,
               aes(x = xstart, xend = xend, y = y_top, yend = y_top),
               inherit.aes = FALSE) +
  # Significance label
  geom_text(data = bracket_df,
            aes(x = xmid, y = y_text, label = annotations),
            size = 5, inherit.aes = FALSE)
p3
```

## IL-10 Expression Violin Plot
```{r Figure_6D_IL10}
# Extract IL10 expression for selected samples
genes_sub <- c(rownames(rld)[grepl("TGFB", rownames(rld))],
               rownames(rld)[grepl("IL10", rownames(rld))],
               "TNF", "IL6", "IL1B", "IL12B", "IL12RB1", "ARG1")
de_mat    <- assay(rld)[genes_sub, c(1:7, 15:29, 38:45)]

my_sample_col <- data.frame(
  Group = rep(c("in vitro nonPMC", "in vitro PMC", "in vitro Mφ", "in vitro PMC Mφ"), c(7, 7, 8, 8))
)
rownames(my_sample_col) <- colnames(de_mat)

de_long <- as.data.frame(t(de_mat)) %>%
  rownames_to_column(var = "Sample") %>%
  pivot_longer(-Sample, names_to = "Feature", values_to = "Expression") %>%
  left_join(rownames_to_column(my_sample_col, var = "Sample"), by = "Sample") %>%
  filter(Feature == "IL10")

de_long$Group <- factor(de_long$Group,
                        levels = c("in vitro nonPMC", "in vitro PMC",
                                   "in vitro Mφ", "in vitro PMC Mφ"))

# Normality check
model <- lm(Expression ~ Group, data = de_long)
ggqqplot(residuals(model))
shapiro_test(residuals(model))

# One-way ANOVA and Tukey HSD
anova <- aov(Expression ~ Group, data = de_long)
summary(anova)
tukey <- TukeyHSD(anova)
print(tukey)

# Build bracket data frame
group_levels <- levels(de_long$Group)
bracket_df <- make_bracket_df(tukey, group_levels, de_long$Expression,
                              bracket_height = 0.002, y_step = 0.01)
bracket_df$y_text <- bracket_df$y_top + 0.01 * bracket_df$y.position

p4 <- ggplot(de_long, aes(x = Group, y = Expression, fill = Group)) +
  geom_violin(width = 0.2, alpha = 1, aes(color = Group)) +
  geom_point(position = position_jitter(width = 0.2), size = 0.5, aes(color = Group)) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 1,
               color = "lightgrey", linewidth = 0.2) +
  scale_fill_manual(values  = my_color_PMC) +
  scale_color_manual(values = my_color_PMC) +
  facet_wrap(~ Feature, scales = "free_y") +
  theme_classic(base_size = 20) +
  labs(x = "", y = "IL10") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text   = element_text(color = "black"),
        axis.title  = element_text(color = "black")) +
  geom_segment(data = bracket_df,
               aes(x = xstart, xend = xstart, y = y.position, yend = y_top),
               inherit.aes = FALSE) +
  geom_segment(data = bracket_df,
               aes(x = xend, xend = xend, y = y.position, yend = y_top),
               inherit.aes = FALSE) +
  geom_segment(data = bracket_df,
               aes(x = xstart, xend = xend, y = y_top, yend = y_top),
               inherit.aes = FALSE) +
  geom_text(data = bracket_df,
            aes(x = xmid, y = y_text, label = annotations),
            size = 5, inherit.aes = FALSE)
p4
```


# Figure 6E: ELISA of TGF-β and IL-10

## Load and Reshape ELISA Data
```{r Figure_6E_load_data}
dat <- data.frame(openxlsx::read.xlsx(
  "Figure 6E ELISAs of TGFb and IL-10.xlsx", sheet = 1
))

dat_long <- dat %>%
  rename(sample = X1) %>%
  pivot_longer(cols = c("Non.PMC", "PMC.1.100"),
               names_to = "Group", values_to = "Expression") %>%
  mutate(
    CellType = case_when(
      endsWith(Group, "Non.PMC")   ~ "in vitro nonPMC",
      endsWith(Group, "PMC.1.100") ~ "in vitro PMC",
      TRUE                          ~ NA_character_
    ),
    Day = case_when(
      startsWith(as.character(Day), "7") ~ "6d",
      startsWith(as.character(Day), "2") ~ "24 hrs",
      TRUE                               ~ NA_character_
    )
  )
dat_long$Gene <- factor(dat_long$Gene, levels = c("TGFb", "IL10"))

# Normality check on ELISA data
model <- lm(Expression ~ Group, data = de_long)
ggqqplot(residuals(model))
shapiro_test(residuals(model))
```

## Plot: Day 1 (24 hrs) — Monocyte vs PMC
```{r Figure_6E_day1}
p5 <- ggplot(dat_long[dat_long$Day == "24 hrs", ],
             aes(x = CellType, y = Expression, fill = CellType)) +
  geom_violin(width = 0.2, alpha = 1, aes(color = CellType)) +
  geom_point(position = position_jitter(width = 0.2), size = 0.5,
             aes(color = Group)) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 1,
               color = "lightgrey", linewidth = 0.2) +
  scale_fill_manual(values  = my_color_PMC) +
  scale_color_manual(values = my_color_PMC) +
  facet_wrap(Day ~ Gene, scales = "free_y") +
  stat_compare_means(
    method      = "t.test",
    label       = "p.format",
    comparisons = list(c("in vitro nonPMC", "in vitro PMC")),
    label.y.npc = "top",
    size        = 5
  ) +
  theme_classic(base_size = 20) +
  labs(x = "", y = "pg/mL") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text   = element_text(color = "black"),
        axis.title  = element_text(color = "black")) +
  NoLegend() +
  scale_y_continuous(expand = c(0, 1000))
p5
```

## Plot: Day 6 (6d) — Mφ vs PMC Mφ
```{r Figure_6E_day6}
dat_long <- dat_long %>%
  mutate(
    CellType_2 = case_when(
      endsWith(CellType, "nonPMC") ~ "in vitro Mφ",
      endsWith(CellType, "PMC")    ~ "in vitro PMC Mφ",
      TRUE                          ~ NA_character_
    )
  )

p6 <- ggplot(dat_long[dat_long$Day == "6d", ],
             aes(x = CellType_2, y = Expression, fill = CellType_2)) +
  geom_violin(width = 0.2, alpha = 1, aes(color = Group)) +
  geom_point(position = position_jitter(width = 0.2), size = 0.5,
             aes(color = Group)) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 1,
               color = "lightgrey", linewidth = 0.2) +
  scale_fill_manual(values  = my_color_PMC) +
  scale_color_manual(values = my_color_PMC) +
  facet_wrap(Day ~ Gene, scales = "free_y") +
  stat_compare_means(
    method          = "t.test",
    label           = "p.format",
    p.adjust.method = "fdr",
    comparisons     = list(c("in vitro Mφ", "in vitro PMC Mφ")),
    label.y.npc     = "top",
    size            = 5
  ) +
  theme_classic(base_size = 20) +
  labs(x = "", y = "pg/mL") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text   = element_text(color = "black"),
        axis.title  = element_text(color = "black")) +
  NoLegend()
p6

# Compute and save t-test p-values for both time points
pval_day1 <- compare_means(
  formula = Expression ~ CellType,
  data    = dat_long %>% filter(Day == "24 hrs"),
  method  = "t.test",
  group.by = "Gene",
  paired  = FALSE
) %>% mutate(Day = "24 hrs")

pval_day6 <- compare_means(
  formula       = Expression ~ CellType_2,
  data          = dat_long %>% filter(Day == "6d"),
  method        = "t.test",
  comparisons   = list(c("in vitro Mφ", "in vitro PMC Mφ")),
  group.by      = "Gene",
  p.adjust.method = "BH"
) %>% mutate(Day = "6d")

pval_df <- bind_rows(pval_day1, pval_day6)
write.csv(pval_df, "Figure 6E T-test with TGFb and IL-10.csv")
```


# Assemble and Save Figure 6 A–E
```{r Figure_6_assemble}
left_plot       <- p1_1 / p1 + plot_layout(heights = c(1, 1))
combined_top    <- left_plot | p2
combined_bottom <- p3 + p4 + p5 + p6 +
  plot_layout(guides = "collect", widths = c(2, 2, 1.5, 4))

combined_full <- combined_top / combined_bottom + plot_layout(heights = c(2, 1.2))

CairoPDF("Figure 6 A-E.pdf", width = 30, height = 16)
print(combined_full)
dev.off()
```


# Supplementary Figure 6A: Volcano Plot (PMC Mφ vs Mφ)
```{r Sup_Figure_6A_volcano}
log2FC_cutoff <- log2(1.5)
padj_cutoff   <- 0.05

need_DEG <- data %>% filter(cluster == "condition_6_vs_4")
need_DEG$significance <- as.factor(
  ifelse(
    need_DEG$padj < padj_cutoff & abs(need_DEG$log2FoldChange) > log2FC_cutoff,
    ifelse(need_DEG$log2FoldChange > log2FC_cutoff, "UP", "DOWN"),
    "NOT"
  )
)

Fig6_supA <- ggplot(data = need_DEG,
                    aes(x = log2FoldChange, y = -log10(padj), color = significance)) +
  geom_point(alpha = 0.5, size = 1) +
  theme_classic(base_size = 16) +
  labs(color = "",
       x     = expression("log"[2] * " (FC)"),
       y     = expression("-log"[10] * " (p-value)")) +
  scale_colour_manual(values = c("#00AFBB", "grey", "#bb0c00")) +
  geom_vline(xintercept = c(-log2FC_cutoff, log2FC_cutoff),
             lty = 4, col = "grey", lwd = 0.8) +
  geom_hline(yintercept = -log10(padj_cutoff),
             lty = 4, col = "grey", lwd = 0.8) +
  theme(
    plot.margin       = unit(c(2, 2, 2, 2), "lines"),
    legend.title      = element_blank(),
    legend.position   = "right",
    axis.text         = element_text(size = 16),
    legend.text       = element_text(size = 16),
    axis.line         = element_line(color = "black", size = 0.8),
    axis.ticks        = element_line(color = "black", size = 0.8),
    axis.ticks.length = unit(0.2, "cm")
  ) +
  xlim(-10, 10) + NoLegend()

# Label top 30 significant genes plus genes of interest
need_DEG <- need_DEG %>%
  mutate(delabel = ifelse(
    significance == "NOT", NA,
    ifelse(
      ID %in% (ID[significance != "NOT"] %>%
                 .[order(padj[significance != "NOT"],
                         -abs(log2FoldChange[significance != "NOT"]))] %>%
                 head(30)),
      ID, NA
    )
  ))

# Force-label key genes of interest
for (g in c("IL10", "TGFBI", "CXCL1", "CCL18", "HIF1A", "IL24", "IL22RA2", "IL15RA")) {
  need_DEG[need_DEG$ID == g, ]$delabel <- g
}

Fig6_supA <- Fig6_supA +
  geom_label_repel(data = need_DEG,
                   aes(log2FoldChange, -log10(padj), label = delabel),
                   max.overlaps = Inf, size = 4)
Fig6_supA
```


# Supplementary Figure 6B: GSEA Bar and Bubble Plots
```{r Sup_Figure_6B_GSEA}
# Pathway enrichment analysis
geneset <- read.gmt("Reactome_human.gmt")
geneset <- geneset[geneset$term %in% top_markers$geneset, ]

# Load DEG data for condition 6 vs 4
group1 <- data %>% filter(cluster == "condition_6_vs_4")

# Extract and rename columns
group1$external_gene_name <- group1$ID
group1$logFC              <- group1$log2FoldChange
group1$FDR                <- group1$padj
group1 <- group1[, c("external_gene_name", "logFC", "FDR")]
group1 <- group1[group1$external_gene_name != "", ]
rownames(group1) <- group1$external_gene_name

# Add up/down regulation column
group1$g1 <- "normal"
group1$g1[group1$logFC >  log2(1) & group1$FDR < 0.05] <- "up"
group1$g1[group1$logFC < -log2(1) & group1$FDR < 0.05] <- "down"
table(group1$g1)

# Get differentially expressed genes and build ranked list
DEGs      <- group1[group1$g1 != "normal", ]
DEGs      <- DEGs[order(DEGs$logFC, decreasing = TRUE), ]
genelist  <- setNames(DEGs$logFC, DEGs$external_gene_name)

# Run GSEA and output all results
egmt <- GSEA(genelist, TERM2GENE = geneset,
             pvalueCutoff = 0.5, pAdjustMethod = "fdr",
             minGSSize = 1, maxGSSize = 500)
dotplot(egmt, showCategory = 10, label_format = 100, color = "pvalue")

# Convert GO term descriptions to factors (ordered for plotting)
dt             <- data.frame(egmt)
dt$Description <- factor(dt$ID, levels = rev(dt$ID))
dt             <- dt %>% filter(NES > 0)

# Enrichment bar plot
Fig6_supB <- ggplot() +
  geom_bar(data = dt,
           aes(x = NES, y = reorder(Description, NES)),
           width = 0.5, fill = my_color_PMC[6], stat = "identity") +
  theme_classic(base_size = 16) +
  scale_x_continuous(expand = c(0, 0)) +
  theme(axis.text.y = element_blank()) +    # Hide y-axis labels (shown as inline text)
  geom_text(data = dt,
            aes(x = 0.01, y = Description, label = Description),
            size = 4.5, hjust = 0) +
  geom_text(data = dt,
            aes(x = 0.01, y = Description, label = core_enrichment),
            color = "grey", size = 4, fontface = "italic", hjust = 0, vjust = 2.3) +
  labs(x = "Normalized Enrichment Score (NES)", y = "Pathways")
Fig6_supB

# Sort pathway names as factor for bubble plot Y-axis
dt       <- dt[order(dt$NES, decreasing = TRUE), ]
dt$ID    <- factor(dt$ID, levels = dt$ID)
dt$xlab  <- 1:nrow(dt)
data_label     <- dt
data_label$col <- "black"

# GSEA bubble plot
Fig6_supBB <- ggplot(data = dt, aes(x = xlab, y = NES)) +
  geom_point(aes(size = setSize, alpha = -log10(pvalue)),
             shape = 21, stroke = 0.7,            # stroke: point border width
             fill = my_color_PMC["in vitro PMC Mφ"], colour = "black") +
  scale_size_continuous(range = c(1, 10)) +
  xlab("") + ylab("Normalized Enrichment Score (NES)") +
  theme_classic(base_size = 16) +
  scale_x_continuous(breaks = seq(0, 50, by = 5)) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.05))) +
  guides(
    size  = guide_legend(title = "Size"),
    alpha = guide_legend(title = expression("-Log"[10] * " (p-value)"))
  ) +
  theme(
    legend.text       = element_text(size = 16),
    axis.line         = element_line(color = "black", size = 0.8),
    axis.ticks        = element_line(color = "black", size = 0.8),
    axis.ticks.length = unit(0.2, "cm"),
    legend.title      = element_blank(),
    axis.text         = element_text(face = "bold")
  ) +
  geom_text_repel(data = dt,
                  aes(x = xlab, y = NES, label = ID),
                  size               = 5,
                  color              = data_label$col,
                  force              = 20,            # Force pushing overlapping labels apart
                  point.padding      = 0.5,           # Minimum distance between label and point
                  min.segment.length = 0,             # Add connector line if length > 0
                  hjust              = 1,             # Align right edge of label to position
                  segment.color      = "grey20",
                  segment.size       = 0.1,           # Connector line width
                  segment.alpha      = 0.8,           # Connector line transparency
                  nudge_y            = -0.1)          # Fine-tune label position along y-axis
Fig6_supBB
```


# Supplementary Figure 6C: scPS Dot Plot — PMC vs nonPMC by Group
```{r Sup_Figure_6C_dotplot}
obj$group <- ifelse(obj$AS_HIV == "ASNN", "HIV-AS-",
             ifelse(obj$AS_HIV == "ASNP", "HIV+AS-",
             ifelse(obj$AS_HIV == "ASPN", "HIV-AS+",
             ifelse(obj$AS_HIV == "ASPP", "HIV+AS+", NA))))

obj$PMC_group <- paste(obj$PMC, obj$group)
Idents(obj) <- "PMC_group"
my_levels   <- c("nonPMC HIV-AS-", "PMC HIV-AS-",
                 "nonPMC HIV+AS-", "PMC HIV+AS-",
                 "nonPMC HIV-AS+", "PMC HIV-AS+",
                 "nonPMC HIV+AS+", "PMC HIV+AS+")
Idents(obj) <- factor(x = Idents(obj), levels = my_levels)

Fig6_supC <- DotPlot(obj, features = features, assay = "scPS", dot.scale = 5) +
  scale_size_area(max_size = 5) +
  coord_flip() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.text.y = element_text(angle = 0,  vjust = 1, hjust = 1)) +
  xlab("") + ylab("") +
  guides(size = guide_legend("Percent Expression")) +
  scale_color_gradientn(colors = colorRampPalette(c("#1F78B4", "white", "#FFED99", "#FF7600"))(10)) +
  theme(
    axis.text         = element_text(size = 16),
    legend.text       = element_text(size = 12),
    legend.title      = element_text(size = 12),
    axis.line         = element_line(color = "black", size = 0.8),
    axis.ticks        = element_line(color = "black", size = 0.8),
    axis.ticks.length = unit(0.2, "cm")
  )
Fig6_supC
```


# Assemble and Save Supplementary Figure 6
```{r Sup_Figure_6_assemble}
combined_sup <- (Fig6_supA + Fig6_supB + plot_layout(widths = c(1, 1.2))) /
  Fig6_supC + plot_layout(heights = c(4, 1))

ggsave(filename = "Figure 6 sup.pdf", plot = combined_sup,
       width = 25, height = 12, bg = "white")

CairoPDF("Figure 6 F.pdf", width = 25, height = 4.5)
Fig6_supC
dev.off()
```
