---
title: "Figure 3: DE Genes and Random Forest ML Analysis"
output: html_document
date: "2024-08-29"
editor_options: 
  chunk_output_type: console
---


# Load Libraries and Set Parameters
```{r setup, message=FALSE, warning=FALSE}
library(readxl)
library(openxlsx)
library(dplyr)
library(tidyverse)
library(reshape2)
library(ggplot2)
library(ggpubr)
library(ggprism)
library(ggVennDiagram)
library(VennDiagram)
library(UpSetR)
library(ComplexHeatmap)
library(scCustomize)
library(Seurat)
library(randomForest)
library(rfPermute)
library(randomForestExplainer)
library(caret)
library(pROC)
library(data.table)
library(vip)
library(doParallel)
library(forcats)

knitr::opts_chunk$set(output.width = 120)

# DE filtering thresholds
p_value      <- 0.05
FC_threshold <- log2(1)
min_threshold <- 0.1

# Path to DE gene CSV files
folder_path <- "DEGs across clusters and groups/"
```


# Figure 3 Venn Diagrams — DE Genes per Cluster Across Groups

## CD14
```{r Figure_3_Venn_CD14}
csv_files <- list.files(path = folder_path, pattern = "\\.csv$", full.names = TRUE)
csv_files <- csv_files[grepl("CD14_", csv_files)]
csv_list  <- lapply(csv_files, read.csv)
extracted_strings <- sub(".*CD14_(.*)\\.csv", "\\1", csv_files)
names(csv_list)   <- extracted_strings
group <- extracted_strings[extracted_strings %in% c("ASNN_vs_ASNP", "ASNN_vs_ASPN", "ASNN_vs_ASPP")]

DE_genes <- list()
for (i in group) {
  csv_list[[i]] <- csv_list[[i]] %>%
    filter(p_val_adj < p_value & abs(avg_log2FC) > FC_threshold) %>%
    filter(pct.1 > min_threshold | pct.2 > min_threshold)
  DE_genes[[i]] <- csv_list[[i]]$X
}

selected_genes <- list(
  ASNN_vs_ASNP = DE_genes$ASNN_vs_ASNP,
  ASNN_vs_ASPN = DE_genes$ASNN_vs_ASPN,
  ASNN_vs_ASPP = DE_genes$ASNN_vs_ASPP
)
selected_genes <- lapply(selected_genes, function(x) x[!grepl("\\.|MT-|^(RPS|RPL)", x)])

p_CD14 <- ggVennDiagram(selected_genes,
  category.names       = names(selected_genes),
  set_color            = rev(my_color_group[names(my_color_group) != "ASNN"]),
  set_size             = 0,
  label                = "both",
  label_color          = "black",
  label_alpha          = 0,
  label_size           = 5,
  label_percent_digit  = 1,
  edge_size            = 1) +
  scale_fill_gradient(low = "white", high = "white") + NoLegend()
p_CD14

genes_df <- list()
genes_df[["CD14"]] <- do.call(rbind, lapply(names(selected_genes), function(name) {
  data.frame(Gene = selected_genes[[name]], List_Name = name, Cluster = "CD14")
}))
```

## CD14-active
```{r Figure_3_Venn_CD14_active}
csv_files <- list.files(path = folder_path, pattern = "\\.csv$", full.names = TRUE)
csv_files <- csv_files[grepl("CD14-active", csv_files)]
csv_list  <- lapply(csv_files, read.csv)
extracted_strings <- sub(".*CD14-active_(.*)\\.csv", "\\1", csv_files)
names(csv_list)   <- extracted_strings
group <- extracted_strings[extracted_strings %in% c("ASNN_vs_ASNP", "ASNN_vs_ASPN", "ASNN_vs_ASPP")]

DE_genes <- list()
for (i in group) {
  csv_list[[i]] <- csv_list[[i]] %>%
    filter(p_val_adj < p_value & abs(avg_log2FC) > FC_threshold) %>%
    filter(pct.1 > min_threshold | pct.2 > min_threshold)
  DE_genes[[i]] <- csv_list[[i]]$X
}

selected_genes <- list(
  ASNN_vs_ASNP = DE_genes$ASNN_vs_ASNP,
  ASNN_vs_ASPN = DE_genes$ASNN_vs_ASPN,
  ASNN_vs_ASPP = DE_genes$ASNN_vs_ASPP
)
selected_genes <- lapply(selected_genes, function(x) x[!grepl("\\.|MT-|^(RPS|RPL)", x)])

p_CD14_active <- ggVennDiagram(selected_genes,
  category.names       = names(selected_genes),
  set_color            = rev(my_color_group[names(my_color_group) != "ASNN"]),
  set_size             = 0,
  label                = "both",
  label_color          = "black",
  label_alpha          = 0,
  label_size           = 5,
  label_percent_digit  = 1,
  edge_size            = 1) +
  scale_fill_gradient(low = "white", high = "white") + NoLegend()
p_CD14_active

genes_df[["CD14-active"]] <- do.call(rbind, lapply(names(selected_genes), function(name) {
  if (length(selected_genes[[name]]) == 0) {
    data.frame(Gene = "", List_Name = name, Cluster = "CD14-active")
  } else {
    data.frame(Gene = selected_genes[[name]], List_Name = name, Cluster = "CD14-active")
  }
}))
```

## CD16
```{r Figure_3_Venn_CD16}
csv_files <- list.files(path = folder_path, pattern = "\\.csv$", full.names = TRUE)
csv_files <- csv_files[grepl("CD16", csv_files)]
csv_list  <- lapply(csv_files, read.csv)
extracted_strings <- sub(".*CD16_(.*)\\.csv", "\\1", csv_files)
names(csv_list)   <- extracted_strings
group <- extracted_strings[extracted_strings %in% c("ASNN_vs_ASNP", "ASNN_vs_ASPN", "ASNN_vs_ASPP")]

DE_genes <- list()
for (i in group) {
  csv_list[[i]] <- csv_list[[i]] %>%
    filter(p_val_adj < p_value & abs(avg_log2FC) > FC_threshold) %>%
    filter(pct.1 > min_threshold | pct.2 > min_threshold)
  DE_genes[[i]] <- csv_list[[i]]$X
}

selected_genes <- list(
  ASNN_vs_ASNP = DE_genes$ASNN_vs_ASNP,
  ASNN_vs_ASPN = DE_genes$ASNN_vs_ASPN,
  ASNN_vs_ASPP = DE_genes$ASNN_vs_ASPP
)
selected_genes <- lapply(selected_genes, function(x) x[!grepl("\\.|MT-|^(RPS|RPL)", x)])

p_CD16 <- ggVennDiagram(selected_genes,
  category.names       = names(selected_genes),
  set_color            = rev(my_color_group[names(my_color_group) != "ASNN"]),
  set_size             = 0,
  label                = "both",
  label_color          = "black",
  label_alpha          = 0,
  label_size           = 5,
  label_percent_digit  = 1,
  edge_size            = 1) +
  scale_fill_gradient(low = "white", high = "white") + NoLegend()
p_CD16

genes_df[["CD16"]] <- do.call(rbind, lapply(names(selected_genes), function(name) {
  if (length(selected_genes[[name]]) == 0) {
    data.frame(Gene = "", List_Name = name, Cluster = "CD16")
  } else {
    data.frame(Gene = selected_genes[[name]], List_Name = name, Cluster = "CD16")
  }
}))
```

## PMC
```{r Figure_3_Venn_PMC}
csv_files <- list.files(path = folder_path, pattern = "\\.csv$", full.names = TRUE)
csv_files <- csv_files[grepl("PMC", csv_files)]
csv_list  <- lapply(csv_files, read.csv)
extracted_strings <- sub(".*PMC_(.*)\\.csv", "\\1", csv_files)
names(csv_list)   <- extracted_strings
group <- extracted_strings[extracted_strings %in% c("ASNN_vs_ASNP", "ASNN_vs_ASPN", "ASNN_vs_ASPP")]

DE_genes <- list()
for (i in group) {
  csv_list[[i]] <- csv_list[[i]] %>%
    filter(p_val_adj < p_value & abs(avg_log2FC) > FC_threshold) %>%
    filter(pct.1 > min_threshold | pct.2 > min_threshold)
  DE_genes[[i]] <- csv_list[[i]]$X
}

selected_genes <- list(
  ASNN_vs_ASNP = DE_genes$ASNN_vs_ASNP,
  ASNN_vs_ASPN = DE_genes$ASNN_vs_ASPN,
  ASNN_vs_ASPP = DE_genes$ASNN_vs_ASPP
)
selected_genes <- lapply(selected_genes, function(x) x[!grepl("\\.|MT-|^(RPS|RPL)", x)])

p_PMC <- ggVennDiagram(selected_genes,
  category.names       = names(selected_genes),
  set_color            = rev(my_color_group[names(my_color_group) != "ASNN"]),
  set_size             = 0,
  label                = "both",
  label_color          = "black",
  label_alpha          = 0,
  label_size           = 5,
  label_percent_digit  = 1,
  edge_size            = 1) +
  scale_fill_gradient(low = "white", high = "white") + NoLegend()
p_PMC

genes_df[["PMC"]] <- do.call(rbind, lapply(names(selected_genes), function(name) {
  if (length(selected_genes[[name]]) == 0) {
    data.frame(Gene = "", List_Name = name, Cluster = "PMC")
  } else {
    data.frame(Gene = selected_genes[[name]], List_Name = name, Cluster = "PMC")
  }
}))
```


# Figure 3A: Venn Diagram Across All Clusters
```{r Figure_3A}
housekeeping_genes <- read_xlsx("Housekeep gene set.xlsx")

dat <- do.call(rbind, genes_df)
dat <- dat[!grepl("^MT-|^MT|\\.|^(RPS|RPL)", dat$Gene), ]
dat <- dat[!dat$Gene %in% housekeeping_genes$Symbol, ]

dat$group <- ifelse(dat$List_Name == "ASNN_vs_ASPN", "HIV-AS+",
             ifelse(dat$List_Name == "ASNN_vs_ASNP", "HIV+AS-",
             ifelse(dat$List_Name == "ASNN_vs_ASPP", "HIV+AS+", NA)))
dat$Cluster <- gsub("_", "/", dat$Cluster)

selected_genes <- split(dat$Gene, dat$Cluster)

Fig3_A <- ggVennDiagram(selected_genes,
  category.names       = names(selected_genes),
  set_color            = c("#FDBF6F", "#FF7F00", "#B15928", "#CCAFCA"),
  set_size             = 5,
  label                = "count",
  label_color          = "black",
  label_alpha          = 0,
  label_size           = 5,
  label_percent_digit  = 1,
  edge_size            = 1) +
  scale_fill_gradient(low = "white", high = "white")
Fig3_A
ggsave(filename = "Figure 3A.pdf", width = 8, height = 5, bg = "white")

# Save unique DE gene list for ML model input
data <- data.frame(Gene = unique(dat$Gene))
write.csv(data, "Figure 3 DEGs for ML.csv")
```


# Upregulated DE Genes Across Clusters (for ML Feature Input)
```{r Upregulated_DEGs_per_cluster}
# Reset thresholds for upregulated-only filter
p_value      <- 0.05
FC_threshold <- log2(1)

# CD14 — upregulated only
csv_files <- list.files(path = folder_path, pattern = "\\.csv$", full.names = TRUE)
csv_files <- csv_files[grepl("CD14_", csv_files) & !grepl("CD14-active", csv_files)]
csv_list  <- lapply(csv_files, read.csv)
extracted_strings <- sub(".*CD14_(.*)\\.csv", "\\1", csv_files)
names(csv_list)   <- extracted_strings
group <- extracted_strings[extracted_strings %in% c("ASNN_vs_ASNP", "ASNN_vs_ASPN", "ASNN_vs_ASPP")]

DE_genes <- list()
for (i in group) {
  csv_list[[i]] <- csv_list[[i]] %>% filter(p_val_adj < p_value & avg_log2FC > FC_threshold)
  DE_genes[[i]] <- csv_list[[i]]$X
}
selected_genes <- list(
  ASNN_vs_ASNP = DE_genes$ASNN_vs_ASNP,
  ASNN_vs_ASPN = DE_genes$ASNN_vs_ASPN,
  ASNN_vs_ASPP = DE_genes$ASNN_vs_ASPP
)
genes_df_up <- list()
genes_df_up[["CD14"]] <- do.call(rbind, lapply(names(selected_genes), function(name) {
  data.frame(Gene = selected_genes[[name]], List_Name = name, Cluster = "CD14")
}))

# CD14-active — upregulated only
csv_files <- list.files(path = folder_path, pattern = "\\.csv$", full.names = TRUE)
csv_files <- csv_files[grepl("CD14-active", csv_files)]
csv_list  <- lapply(csv_files, read.csv)
extracted_strings <- sub(".*CD14-active_(.*)\\.csv", "\\1", csv_files)
names(csv_list)   <- extracted_strings
group <- extracted_strings[extracted_strings %in% c("ASNN_vs_ASNP", "ASNN_vs_ASPN", "ASNN_vs_ASPP")]

DE_genes <- list()
for (i in group) {
  csv_list[[i]] <- csv_list[[i]] %>% filter(p_val_adj < p_value & avg_log2FC > FC_threshold)
  DE_genes[[i]] <- csv_list[[i]]$X
}
selected_genes <- list(
  ASNN_vs_ASNP = DE_genes$ASNN_vs_ASNP,
  ASNN_vs_ASPN = DE_genes$ASNN_vs_ASPN,
  ASNN_vs_ASPP = DE_genes$ASNN_vs_ASPP
)
genes_df_up[["CD14-active"]] <- do.call(rbind, lapply(names(selected_genes), function(name) {
  data.frame(Gene = selected_genes[[name]], List_Name = name, Cluster = "CD14-active")
}))

# CD16 — upregulated only
csv_files <- list.files(path = folder_path, pattern = "\\.csv$", full.names = TRUE)
csv_files <- csv_files[grepl("CD16", csv_files)]
csv_list  <- lapply(csv_files, read.csv)
extracted_strings <- sub(".*CD16_(.*)\\.csv", "\\1", csv_files)
names(csv_list)   <- extracted_strings
group <- extracted_strings[extracted_strings %in% c("ASNN_vs_ASNP", "ASNN_vs_ASPN", "ASNN_vs_ASPP")]

DE_genes <- list()
for (i in group) {
  csv_list[[i]] <- csv_list[[i]] %>% filter(p_val_adj < p_value & avg_log2FC > FC_threshold)
  DE_genes[[i]] <- csv_list[[i]]$X
}
selected_genes <- list(
  ASNN_vs_ASNP = DE_genes$ASNN_vs_ASNP,
  ASNN_vs_ASPN = DE_genes$ASNN_vs_ASPN,
  ASNN_vs_ASPP = DE_genes$ASNN_vs_ASPP
)
genes_df_up[["CD16"]] <- do.call(rbind, lapply(names(selected_genes), function(name) {
  data.frame(Gene = selected_genes[[name]], List_Name = name, Cluster = "CD16")
}))

# PMC — upregulated only
csv_files <- list.files(path = folder_path, pattern = "\\.csv$", full.names = TRUE)
csv_files <- csv_files[grepl("PMC", csv_files)]
csv_list  <- lapply(csv_files, read.csv)
extracted_strings <- sub(".*PMC_(.*)\\.csv", "\\1", csv_files)
names(csv_list)   <- extracted_strings
group <- extracted_strings[extracted_strings %in% c("ASNN_vs_ASNP", "ASNN_vs_ASPN", "ASNN_vs_ASPP")]

DE_genes <- list()
for (i in group) {
  csv_list[[i]] <- csv_list[[i]] %>% filter(p_val_adj < p_value & avg_log2FC > FC_threshold)
  DE_genes[[i]] <- csv_list[[i]]$X
}
selected_genes <- list(
  ASNN_vs_ASNP = DE_genes$ASNN_vs_ASNP,
  ASNN_vs_ASPN = DE_genes$ASNN_vs_ASPN,
  ASNN_vs_ASPP = DE_genes$ASNN_vs_ASPP
)
genes_df_up[["PMC"]] <- do.call(rbind, lapply(names(selected_genes), function(name) {
  data.frame(Gene = selected_genes[[name]], List_Name = name, Cluster = "PMC")
}))

# Combine, filter, and save upregulated DEG table
dat_up <- do.call(rbind, genes_df_up)
dat_up <- dat_up[!grepl("^MT-|^MT|\\.|^(RPS|RPL)", dat_up$Gene), ]
dat_up <- dat_up[!dat_up$Gene %in% housekeeping_genes$Symbol, ]
dat_up <- dat_up[dat_up$Gene %in% Genes_ML$Gene, ]
dat_up$group <- ifelse(dat_up$List_Name == "ASNN_vs_ASPN", "HIV-AS+",
               ifelse(dat_up$List_Name == "ASNN_vs_ASNP", "HIV+AS-",
               ifelse(dat_up$List_Name == "ASNN_vs_ASPP", "HIV+AS+", NA)))
dat_up$Cluster <- gsub("_", "/", dat_up$Cluster)
write.csv(dat_up, "Figure 3 upregulated DEGs.csv")
```


# Random Forest ML Analysis

## Step 1: Prepare Expression and Metadata
```{r RF_prepare_data}
Genes_ML  <- read.csv("Figure 3 DEGs for ML.csv", header = TRUE)
Genes_ML  <- Genes_ML[!grepl("^MT-|^MT|\\.|^(RPS|RPL)", Genes_ML$Gene), ]

# Exclude intermediate clusters
small_obj <- subset(x = obj,
                    idents = c("Intermediate", "Intermediate-metabolic", "Intermediate-interferon"),
                    invert = TRUE)
table(small_obj$cluster_num, small_obj$AS_HIV)

SCT_data  <- small_obj[["SCT"]]@data[Genes_ML$Gene, ]
meta_data <- small_obj@meta.data
meta_data$cell_id <- rownames(meta_data)
group_table <- unique(meta_data[, c("cluster_num", "AS_HIV")])
```

## Step 2: Define Sampling and Averaging Function
```{r RF_sampling_function}
set.seed(123)
sample_and_average <- function(cluster, hiv_group, expr_data, meta_data, n_cells = 100) {
  cells_in_group <- meta_data %>%
    filter(cluster_num == cluster, AS_HIV == hiv_group) %>%
    pull(cell_id)
  if (length(cells_in_group) < n_cells) return(NULL)
  sampled_cells <- sample(cells_in_group, n_cells)
  averaged_expr <- rowMeans(expr_data[, sampled_cells, drop = FALSE])
  return(averaged_expr)
}
```

## Step 3: Generate Training Data by Bootstrap Sampling
```{r RF_generate_training_data}
Genes <- rownames(SCT_data)
result_list <- list()

for (i in 1:nrow(group_table)) {
  cluster   <- group_table$cluster_num[i]
  hiv_group <- group_table$AS_HIV[i]
  group_name <- paste(cluster, hiv_group, sep = "_")
  cat("Processing group:", group_name, "\n")
  avg_matrix <- matrix(NA, nrow = length(Genes), ncol = 100)
  rownames(avg_matrix) <- Genes
  for (rep in 1:100) {
    set.seed(123 + i * 1000 + rep)
    avg_matrix[, rep] <- sample_and_average(
      cluster   = cluster,
      hiv_group = hiv_group,
      expr_data = SCT_data,
      meta_data = meta_data,
      n_cells   = 100
    )
  }
  result_list[[group_name]] <- avg_matrix
}

df_list <- list()
for (group_name in names(result_list)) {
  avg_matrix         <- as.data.frame(t(result_list[[group_name]]))
  colnames(avg_matrix) <- gsub("-", "_", colnames(avg_matrix))
  avg_matrix$status  <- gsub("-", "_", group_name)
  df_list[[group_name]] <- avg_matrix
}

data <- bind_rows(df_list)
data <- data %>% mutate(across(-ncol(data), as.numeric))
data$status <- as.factor(data$status)
```

## Step 4: Split into Training and Validation Sets
```{r RF_split_data}
set.seed(123)
train_index <- createDataPartition(y = data$status, p = 0.7, list = FALSE)
train_data  <- data[train_index, ]
val_data    <- data[-train_index, ]
```

## Step 5: Train Initial Random Forest Model
```{r RF_initial_model}
train_x <- train_data[, -ncol(data)]
train_y <- train_data[, ncol(data)]

set.seed(123)
rf_model0 <- randomForest(status ~ .,
                          data      = train_data,
                          importance = TRUE,
                          proximity  = TRUE)
rf_model0

# Plot OOB error vs number of trees
plot(rf_model0, main = "OOB Error vs Number of Trees")
n <- ncol(rf_model0$err.rate)
legend("topright",
       legend = colnames(rf_model0$err.rate),
       lty = 1:n, col = 1:n)

# Check number of trees at minimum overall OOB error
which.min(rf_model0$err.rate[, 1])

# Visualize tree node counts
hist(treesize(rf_model0), main = "Number of Nodes per Tree", col = "red")

# Training set confusion matrix
predict_initial <- predict(rf_model0, newdata = train_data)
matrix0 <- table(train_y, predict_initial)
matrix0 <- round(matrix0 / rowSums(matrix0), 2)
ggplot(data = as.data.frame(matrix0),
       aes(x = predict_initial, y = train_y, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = Freq), vjust = 1) +
  scale_fill_gradient(low = "white", high = "red") +
  theme_minimal() +
  labs(title = "Confusion Matrix (Training Set)", x = "Predicted", y = "Actual")
```

## Step 6: Hyperparameter Tuning
```{r RF_hyperparameter_tuning}
# Determine approximate number of features for mtry search
sqrt(ncol(train_x))

# Method: use caret with 5-fold cross-validation to tune mtry
set.seed(123)
trainControl <- trainControl(
  method          = "cv",
  summaryFunction = multiClassSummary,
  number          = 5,
  classProbs      = TRUE,
  savePredictions = TRUE
)
cl <- makeCluster(detectCores() - 1)
registerDoParallel(cl)
levels(train_data$status) <- gsub("\\/", "_", levels(train_data$status))
rfFit <- train(status ~ .,
               data      = train_data,
               method    = "rf",
               ntree     = 100,
               trControl = trainControl)
stopCluster(cl)
registerDoSEQ()
gc()
rfFit
# The final value used for the model was mtry = 45

# Tune ntree by fixing best mtry and comparing accuracy across ntree values
set.seed(123)
ctrl      <- trainControl(method = "cv", number = 5)
bestgrid  <- expand.grid(mtry = 45)
modellist <- list()
ntree_list <- c(50, 100, 200, 300)
for (ntree in ntree_list) {
  cat("Training model with ntree =", ntree, "\n")
  fit <- train(x          = train_x,
               y          = train_y,
               method     = "rf",
               metric     = "Accuracy",
               tuneGrid   = bestgrid,
               trControl  = ctrl,
               ntree      = ntree)
  modellist[[as.character(ntree)]] <- fit
}
results <- resamples(modellist)
summary(results)
# Based on accuracy, ntree = 100 is optimal
```

## Step 7: Final Random Forest Model and Variable Importance
```{r RF_final_model}
# Retrain with best hyperparameters (mtry = 45, ntree = 200)
set.seed(123)
rf_model2 <- randomForest(x         = train_x,
                          y         = train_y,
                          mtry      = 45,
                          ntree     = 200,
                          nodesize  = 1,
                          importance = TRUE,
                          proximity  = TRUE)
print(rf_model2)
plot(rf_model2, lwd = 2, main = "OOB Error")

n <- 30
varImpPlot(rf_model2, main = "Variable Importance (Accuracy)", type = 1,
           n.var = n, scale = TRUE, cex = 1)
varImpPlot(rf_model2, main = "Variable Importance (Gini)", type = 2,
           n.var = 30, scale = TRUE, cex = 1)
```

## Step 8: Variable Importance Bar Plot
```{r RF_variable_importance}
importance    <- varImpPlot(rf_model2, main = "Variable Importance (Accuracy)",
                            type = 1, n.var = n, scale = TRUE, cex = 1)
importance_df <- as.data.frame(importance)
importance_df$variable <- rownames(importance_df)
importance_df <- importance_df[order(-importance_df$MeanDecreaseAccuracy), ]

# Set gradient color range
color_range <- c("#C8EBF6", "#154778")

# Horizontal bar chart of top variable importances
ggplot(importance_df[1:n, ],
       aes(x = reorder(variable, MeanDecreaseAccuracy),
           y = MeanDecreaseAccuracy,
           fill  = MeanDecreaseAccuracy,
           label = round(MeanDecreaseAccuracy, 2))) +
  geom_bar(stat = "identity") +
  geom_text(size = 3, position = position_stack(vjust = 0.5), color = "black") +
  scale_fill_gradient(low = color_range[1], high = color_range[2]) +
  theme(
    plot.title  = element_text(hjust = 0.5, size = 16),
    axis.text   = element_text(size = 12),
    axis.title  = element_text(size = 15)
  ) +
  labs(title = "", x = "", y = "Mean Decrease Accuracy") +
  coord_flip() + theme_classic()
```

## Step 9: MDS Plot
```{r RF_MDS_plot}
MDSplot(rf_model2, train_data$status,
        main = "Multi-Dimensional Scaling Plot of Proximity Matrix",
        pch  = 19)
legend("right",
       legend  = levels(train_data$status),
       col     = 1:length(levels(train_data$status)),
       pch     = 19,
       cex     = 0.8,
       xpd     = TRUE,
       inset   = -1)
```

## Step 10: Apply Model to Validation Set
```{r RF_validation}
val_x <- val_data[, -ncol(data)]
val_y <- val_data[, ncol(data)]

# Predict on validation set
predict_val <- predict(rf_model2, newdata = val_x)
matrix0     <- table(val_y, predict_val)
matrix0     <- round(matrix0 / rowSums(matrix0), 2)

ggplot(data = as.data.frame(matrix0),
       aes(x = predict_val, y = val_y, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = Freq), vjust = 1) +
  scale_fill_gradient(low = "white", high = "red") +
  theme_minimal() +
  labs(title = "Confusion Matrix (Validation Set)", x = "Predicted", y = "Actual")
```

## Step 11: ROC Curves (Validation Set)
```{r RF_ROC_curves}
val_predprob  <- predict(rf_model2, newdata = val_data, type = "prob")
val_classes   <- colnames(val_predprob)
roc_list_val  <- list()
auc_results_val <- data.frame(Class = character(), AUC = numeric(), stringsAsFactors = FALSE)

for (cls in val_classes) {
  response_bin <- ifelse(val_data$status == cls, 1, 0)
  roc_obj      <- roc(response = response_bin, predictor = val_predprob[, cls])
  roc_df       <- data.frame(
    Specificity = 1 - roc_obj$specificities,
    Sensitivity = roc_obj$sensitivities,
    Class = cls
  )
  roc_list_val[[cls]] <- roc_df
  auc_results_val <- rbind(auc_results_val,
                           data.frame(Class = cls, AUC = as.numeric(auc(roc_obj))))
}

# Merge all ROC curve data
roc_data_val <- do.call(rbind, roc_list_val)

# Multi-class ROC plot
Fig3_supA <- ggplot(roc_data_val, aes(x = Specificity, y = Sensitivity, color = Class)) +
  geom_line(size = 1.2) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  theme_classic() +
  labs(title = "Multi-class ROC Curves (Validation Set)",
       x = "1 - Specificity", y = "Sensitivity")
Fig3_supA
ggsave(filename = "Figure 3 Sup A.pdf", width = 8, height = 6, bg = "white")
```

## Step 12: Calibration Curves (Validation Set)
```{r RF_calibration}
calibration_data_val <- data.frame()
true_classes_val     <- as.character(val_data$status)

for (cls in val_classes) {
  response_bin   <- ifelse(true_classes_val == cls, 1, 0)
  predictor_prob <- val_predprob[, cls]
  cuts           <- quantile(predictor_prob, probs = seq(0, 1, 0.1), na.rm = TRUE)
  cut_groups     <- cut(predictor_prob, breaks = unique(cuts), include.lowest = TRUE)
  mean_pred      <- tapply(predictor_prob, cut_groups, mean)
  mean_obs       <- tapply(response_bin,   cut_groups, mean)
  df             <- data.frame(Predicted = mean_pred, Observed = mean_obs, Class = cls)
  calibration_data_val <- rbind(calibration_data_val, df)
}
calibration_data_val <- calibration_data_val[complete.cases(calibration_data_val), ]

ggplot(calibration_data_val, aes(x = Predicted, y = Observed, color = Class)) +
  geom_point(size = 2) +
  geom_line(size = 1) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  theme_minimal() +
  labs(title = "Calibration Curves (Validation Set)",
       x = "Mean Predicted Probability",
       y = "Observed Proportion") +
  theme(legend.title = element_blank())
```

## Step 13: Random Forest Model Explainer
```{r RF_explainer}
library(randomForestExplainer)

# Minimum depth distribution
min_depth_frame <- min_depth_distribution(rf_model2)
head(min_depth_frame, n = 10)
plot_min_depth_distribution(min_depth_frame)

# Compute multiple variable importance metrics simultaneously
importance_forest <- measure_importance(rf_model2)

# Visualize mean_min_depth vs times_a_root, with point size = no_of_nodes (3 metrics)
Fig3_supB <- plot_multi_way_importance(importance_forest,
                                       size_measure  = "no_of_nodes",
                                       no_of_labels  = 100) + ggtitle("")
Fig3_supB

# Visualize accuracy_decrease vs gini_decrease with point size = p_value
Fig3_supC <- plot_multi_way_importance(importance_forest,
                                       x_measure    = "accuracy_decrease",
                                       y_measure    = "gini_decrease",
                                       size_measure = "p_value",
                                       no_of_labels = 100) + ggtitle("")
Fig3_supC

Fig3_supABC <- Fig3_supA + Fig3_supB + Fig3_supC
ggsave(filename = "Figure 3 Sup A-C.pdf", width = 20, height = 6, bg = "white")
```

## Step 14: rfPermute — Permutation-Based Significance Testing
```{r RF_permute}
pmt2 <- rfPermute(status ~ .,
                  data      = train_data,
                  ntree     = 200,
                  mtry      = 45,
                  importance = TRUE,
                  num.rep   = 100,
                  num.cores = parallel::detectCores() - 1)
saveRDS(pmt2, file = "RandomForest Permute model.rds")

# Extract and merge permutation p-values
pvalues2           <- data.frame(pmt2$pval[, , 2][, (ncol(pmt2$pval) - 1):ncol(pmt2$pval)])
colnames(pvalues2) <- c("Accuracy.p", "Gini.p")
pvalues2$variable  <- rownames(pvalues2)
rf_result2         <- merge(importance_forest, pvalues2, by = "variable")

# Add significance labels
rf_result2$Gini.p_sig <- ifelse(rf_result2$Gini.p <= 0.001, "***",
                          ifelse(rf_result2$Gini.p <= 0.01, "**",
                          ifelse(rf_result2$Gini.p <= 0.05, "*", "")))
rf_result2$Accuracy.p_sig <- ifelse(rf_result2$Accuracy.p <= 0.001, "***",
                              ifelse(rf_result2$Accuracy.p <= 0.01, "**",
                              ifelse(rf_result2$Accuracy.p <= 0.05, "*", "")))
```

## Step 15: Filter Top Features and Plot (Supplementary Figure 3D)
```{r RF_top_features}
# Filter genes by multi-criteria importance thresholds
rf_subset <- subset(rf_result2,
  (times_a_root > quantile(times_a_root, 0.75, na.rm = TRUE) &
     mean_min_depth < quantile(mean_min_depth, 0.25, na.rm = TRUE)) |
  (gini_decrease > quantile(gini_decrease, 0.75, na.rm = TRUE) &
     accuracy_decrease > quantile(accuracy_decrease, 0.75, na.rm = TRUE)) |
  (Accuracy.p < 0.05 & Gini.p < 0.05)
)
rf_subset <- rf_subset[order(-rf_subset$accuracy_decrease, -rf_subset$gini_decrease), ]
rf_subset$variable <- factor(rf_subset$variable, levels = rf_subset$variable)
rf_subset <- rf_subset[rf_subset$variable != "XIST", ]

color_range <- c("#C8EBF6", "#154778")

Fig3_supD <- ggplot(rf_subset[1:100, ],
                    aes(x = variable, y = accuracy_decrease, fill = accuracy_decrease)) +
  geom_bar(stat = "identity", width = 0.8) +
  geom_text(aes(label = Accuracy.p_sig, y = accuracy_decrease), size = 5) +
  scale_fill_gradient(low = color_range[1], high = color_range[2]) +
  labs(y = "Accuracy Decrease Score", x = "") +
  theme_classic() +
  theme(
    axis.text.y  = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid   = element_blank(),
    axis.text.x  = element_text(size = 10, face = "bold", angle = 90, hjust = 1)
  )

Fig3_supABC / Fig3_supD
ggsave(filename = "Figure 3 Sup A-D.pdf", width = 18, height = 10, bg = "white")
```

## Step 16: Select Top 100 ML Genes
```{r RF_top100_genes}
data <- read.csv("Figure 3 RandomForest permute model importance score filtered.csv", header = TRUE)
data_1 <- data %>%
  arrange(desc(accuracy_decrease)) %>%
  slice_head(n = 100)
```


# Figure 3B: UpSet Plot — ML Top Genes vs Cluster-Specific Upregulated DEGs
```{r Figure_3B}
dat <- read.csv("Figure 3 upregulated genes.csv", header = TRUE)
dat <- dat[dat$Gene %in% data_1$variable, ]
dat$group <- factor(dat$group, levels = c("HIV+AS-", "HIV-AS+", "HIV+AS+"))
genes_df  <- split(dat$Gene, dat$Cluster)

# Intersect ML top genes with cluster-specific upregulated DEGs
DEgenes <- list(
  CD14       = intersect(data_1$variable, genes_df[["CD14"]]),
  CD14_active = intersect(data_1$variable, genes_df[["CD14-active"]]),
  CD16       = intersect(data_1$variable, genes_df[["CD16"]]),
  PMC        = intersect(data_1$variable, genes_df[["PMC"]])
)

# Color settings for UpSet plot
main_bar_col <- "violetred4"
matrix_col   <- "black"
shade_col    <- "grey"

Venn_data <- fromList(DEgenes)
Fig3_B <- upset(Venn_data,
  sets          = c("CD14", "CD14_active", "CD16", "PMC"),
  sets.bar.color = c("#FDBF6F", "#FF7F00", "#B15928", "#CCAFCA"),
  keep.order    = TRUE,
  order.by      = "degree",
  point.size    = 3,
  line.size     = 1,
  number.angles = 0,
  text.scale    = 2,
  main.bar.color = main_bar_col,
  matrix.color  = matrix_col,
  shade.color   = shade_col,
  sets.x.label  = "Size"
)
pdf(file = "Figure 3B.pdf", width = 8, height = 5)
Fig3_B
dev.off()
```

## Check Shared Genes Across Clusters
```{r check_shared_genes}
# Genes shared across all clusters
Reduce(intersect, DEgenes)

# Count how many clusters each gene appears in
gene_counts    <- table(unlist(DEgenes))
shared_genes_4 <- names(gene_counts[gene_counts == 4])
shared_genes_3 <- names(gene_counts[gene_counts == 3])
shared_genes_2 <- names(gene_counts[gene_counts == 2])
shared_genes_1 <- names(gene_counts[gene_counts == 1])
shared_genes_4
shared_genes_3
shared_genes_2
shared_genes_1

# Identify genes unique to PMC
ClusterX        <- "PMC"
other_clusters  <- setdiff(names(DEgenes), ClusterX)
other_genes     <- unlist(DEgenes[other_clusters])
unique_to_ClusterX <- setdiff(DEgenes[[ClusterX]], other_genes)
unique_to_ClusterX
```


# Figure 3C–F: Clustered Dot Plots per Cluster
```{r Figure_3C-F}
# Add group labels to Seurat object
obj$group <- ifelse(obj$AS_HIV == "ASNN", "HIV-AS-",
             ifelse(obj$AS_HIV == "ASNP", "HIV+AS-",
             ifelse(obj$AS_HIV == "ASPN", "HIV-AS+",
             ifelse(obj$AS_HIV == "ASPP", "HIV+AS+", NA))))
my_levels <- c("HIV-AS-", "HIV+AS-", "HIV-AS+", "HIV+AS+")

Idents(obj) <- "cluster_num"
plots <- list()

# CD14
obj_downsample <- subset(x = obj, idents = "CD14")
Idents(obj_downsample) <- factor(obj_downsample$group, levels = my_levels)
plots[["CD14"]] <- Clustered_DotPlot(
  seurat_object    = obj_downsample,
  assay            = "SCT",
  cluster_ident    = FALSE,
  x_lab_rotate     = 90,
  row_names_side   = "left",
  row_label_size   = 12,
  column_label_size = 12,
  flip             = FALSE,
  colors_use_exp   = colorRampPalette(c("#1F78B4", "white", "#FFED99", "#FF7600"))(10),
  features         = DEgenes$CD14,
  k                = 4,
  colors_use_idents = rev(my_color_group),
  seed             = 12345
)[[2]]

# CD14-active
obj_downsample <- subset(x = obj, idents = "CD14-active")
Idents(obj_downsample) <- factor(obj_downsample$group, levels = my_levels)
plots[["CD14-active"]] <- Clustered_DotPlot(
  seurat_object    = obj_downsample,
  cluster_ident    = FALSE,
  x_lab_rotate     = 90,
  row_names_side   = "left",
  row_label_size   = 12,
  column_label_size = 12,
  flip             = FALSE,
  colors_use_exp   = colorRampPalette(c("#1F78B4", "white", "#FFED99", "#FF7600"))(10),
  features         = DEgenes$CD14_active,
  k                = 4,
  colors_use_idents = rev(my_color_group),
  seed             = 12345
)[[2]]

# CD16
obj_downsample <- subset(x = obj, idents = "CD16")
Idents(obj_downsample) <- factor(obj_downsample$group, levels = my_levels)
plots[["CD16"]] <- Clustered_DotPlot(
  seurat_object    = obj_downsample,
  cluster_ident    = FALSE,
  x_lab_rotate     = 90,
  row_names_side   = "left",
  row_label_size   = 12,
  column_label_size = 12,
  flip             = FALSE,
  colors_use_exp   = colorRampPalette(c("#1F78B4", "white", "#FFED99", "#FF7600"))(10),
  features         = DEgenes$CD16,
  k                = 4,
  colors_use_idents = rev(my_color_group),
  seed             = 12345
)[[2]]

# PMC
obj_downsample <- subset(x = obj, idents = "PMC")
Idents(obj_downsample) <- factor(obj_downsample$group, levels = my_levels)
plots[["PMC"]] <- Clustered_DotPlot(
  seurat_object    = obj_downsample,
  cluster_ident    = FALSE,
  x_lab_rotate     = 90,
  row_names_side   = "left",
  row_label_size   = 12,
  column_label_size = 12,
  flip             = FALSE,
  colors_use_exp   = colorRampPalette(c("#1F78B4", "white", "#FFED99", "#FF7600"))(10),
  features         = DEgenes$PMC,
  k                = 4,
  colors_use_idents = rev(my_color_group),
  seed             = 12345
)[[2]]

# Arrange all four plots in a 2x2 grid using ComplexHeatmap viewports
grid.newpage()
pushViewport(viewport(layout = grid.layout(nrow = 2, ncol = 2)))
pushViewport(viewport(layout.pos.row = 1, layout.pos.col = 1)); draw(plots[["CD14"]],       newpage = FALSE); popViewport()
pushViewport(viewport(layout.pos.row = 2, layout.pos.col = 1)); draw(plots[["CD14-active"]], newpage = FALSE); popViewport()
pushViewport(viewport(layout.pos.row = 1, layout.pos.col = 2)); draw(plots[["CD16"]],        newpage = FALSE); popViewport()
pushViewport(viewport(layout.pos.row = 2, layout.pos.col = 2)); draw(plots[["PMC"]],         newpage = FALSE); popViewport()
# Save as portrait 18 x 15 inches
```


# Step 17: Cross-Validation Recursive Feature Elimination
```{r RF_cross_validation}
set.seed(123)
train.rfcv <- rfcv(trainx   = train_data[, -ncol(data)],
                   trainy   = train_data$status,
                   cv.fold  = 5,
                   recursive = TRUE)
train.rfcv$n.var
train.rfcv$error.cv
with(train.rfcv, plot(n.var, error.cv, type = "o", lwd = 2))

# Repeat cross-validation 5 times and plot smoothed error curve
train.re.rfcv           <- data.frame(sapply(train.re.rfcv, "[[", "error.cv"))
train.re.rfcv$variables <- rownames(train.re.rfcv)
train.re.rfcv           <- reshape2::melt(train.re.rfcv, id = "variables")
train.re.rfcv$variables <- as.numeric(as.character(train.re.rfcv$variables))

plot.rfcv <- ggplot(train.re.rfcv, aes(variables, value)) +
  geom_smooth(se = FALSE) +
  theme(
    panel.grid       = element_blank(),
    panel.background = element_rect(color = "black", fill = "transparent")
  ) +
  labs(x = "Number of Variables", y = "Cross-Validation Error")
plot.rfcv
```


# Supplementary Figure 3E–F: ML Gene Features by Group and Cluster
```{r Sup_Figure_3E-F}
dat <- read.csv("Figure 3 upregulated DEGs.csv", header = TRUE)
dat <- dat[dat$Gene %in% data_1$variable, ]
dat$group <- factor(dat$group, levels = c("HIV+AS-", "HIV-AS+", "HIV+AS+"))

# Supplementary Figure 3E: Venn diagram by group
selected_genes <- split(dat$Gene, dat$group)
Fig3_Sup_E <- ggVennDiagram(selected_genes,
  category.names       = names(selected_genes),
  set_color            = c("#A6CEE3", "#ffa725", "#ff6b6d"),
  set_size             = 0,
  label                = "count",
  label_color          = "black",
  label_alpha          = 0,
  label_size           = 5,
  label_percent_digit  = 1,
  edge_size            = 1) +
  scale_fill_gradient(low = "white", high = "white")
Fig3_Sup_E
ggsave(filename = "Figure 3 Sup E.pdf", width = 3, height = 3, bg = "white")

# Supplementary Figure 3F: UpSet plot by cluster and group
dat$group <- paste0(dat$Cluster, " ", dat$group)
selected_genes <- split(dat$Gene, dat$group)
Venn_data      <- fromList(selected_genes)

Fig3_Sup_F <- upset(Venn_data,
  sets            = names(selected_genes),
  keep.order      = TRUE,
  order.by        = "freq",
  point.size      = 3,
  line.size       = 1,
  number.angles   = 0,
  text.scale      = 2,
  nintersects     = NA,
  main.bar.color  = "black",
  sets.bar.color  = rep(c("#FDBF6F", "#FF7F00", "#B15928", "#CCAFCA"), each = 3),
  matrix.color    = "black",
  shade.color     = "grey",
  sets.x.label    = ""
)
pdf(file = "Figure 3 Sup F.pdf", width = 25, height = 10)
Fig3_Sup_F
dev.off()
```
